---
title: "Predicting soccer match scores"
author: "[Henrique Ap. Laureano](http://www.leg.ufpr.br/~henrique)"
date: "*Last modification on* `r Sys.time()`"
include-before:
  - '\(\newcommand{\bm}[1]{\boldsymbol{\mathbf{#1}}}\)'
output:
  html_document:
    toc: true
    toc_float: true
    css: font_size.css
    code_folding: show
---

```{r, echo=FALSE}
htmltools::img(src = knitr::image_uri("logo_leg.jpg"),
               alt = 'logo',
               style = 'position:absolute; top:0; right:0;
                        padding:10px; width:200px;')
options(width = 100)
library(knitr)
opts_chunk$set(fig.path = "iBagens/",
               cache = TRUE,
               cache.path = "cache/",
               warning = FALSE,
               comment = NA,
               prompt = FALSE)
```

***

# How?

[INLA](http://www.r-inla.org/).

# Show me!

## Functions

All the build functions to be used in this analysis can be found
[HERE](functions.R).

```{r, cache=FALSE}
source("functions.R")
```

## Data

Premier League results for 2018-19 season.

Data from
[HERE](https://www.soccer24.com/england/premier-league-2018-2019/results/).

```{r}
dat <- read.table("~/Dropbox/NORONHA/PremierLeague-2018-19.txt",
                  col.names = c("round", "home", "away", "x", "y"))

teams <- as.character(sort(unique(dat$home)))
n_teams <- length(teams)
n_matches <- nrow(dat) ; n_rounds <- max(dat$round)

n_table <- matrix(c(n_teams, n_matches, n_rounds))
dimnames(n_table) <- list(c("Teams", "Matches", "Round"), "Number")
n_table
```

The first row

```{r}
dat[1, ]
```

need to become something like:

> TeamA | TeamB | round | home-adv | num goals scored by A vs B in match k

```{r}
dat_model <- data.frame()
for(i in 1:n_matches){
    dat_model = rbind(
        dat_model,
        c(dat$home[i], dat$away[i], dat$round[i],  1, dat$x[i]),
        c(dat$away[i], dat$home[i], dat$round[i], -1, dat$y[i]))
}
names(dat_model) = c("TeamA", "TeamB", "round", "home.adv", "score")
dat_model <-
    dat_model %>%
    as_tibble() %>%
    mutate(round_home = round, round_away = round, round = NULL)

datatable(dat_model, options = list(pageLength = 6),
          class = 'cell-border stripe', rownames = FALSE)
```

## Models

> A Poisson mixed model with one fixed effect and two random effects. As
> fixed effect, a binary variable indicating a possible `home effect`.
> As random effects we have the `team performances/strengths` varying
> along the time, i.e., the 38 league rounds.

\[
 \begin{align}
  x_{i} &\sim \text{Poisson}
  \left(\lambda_{x_{i}} =
        \exp\{\mu_{H} + s_{A_{i}}^{t} - s_{B_{i}}^{t}\}
  \right)\\
  \log \lambda_{x_{i}} &= \mu_{H} + s_{A_{i}}^{t} - s_{B_{i}}^{t},
  \quad i = 1, 3, 5, \dots, \quad t = 1, \dots, 38,\\
   y_{i} &\sim \text{Poisson}
  \left(\lambda_{y_{i}} =
        \exp\{-\mu_{H} + s_{B_{i}}^{t} - s_{A_{i}}^{t}\}
  \right)\\
  \log \lambda_{y_{i}} &= -\mu_{H} + s_{B_{i}}^{t} - s_{A_{i}}^{t},
  \quad i = 2, 4, 6, \dots, \quad t = 1, \dots, 38,
 \end{align}
\]

> To describe those random temporal effects we try here three different
> approachs: a random walk of 1st order (`RW1`); a autoregressive
> process of 1st order (`AR1`); and a one dimensional Matern model where
> we construct a mesh on the time points (rounds) and then apply a joint
> PC prior on the practical range and marginal standard deviation
> parameters of the SPDE model (`Mat`). In the figures below these
> models are named as `A`, `B` and `C`, respectively.

```{r}
w_roundA = rep( 1, times = nrow(dat_model))
w_roundB = rep(-1, times = nrow(dat_model))
```

### RW1 model

```{r}
formula_rw1 <- score ~ -1 + home.adv +
    f(round_home, w_roundA, model = "rw1",
      hyper = list(prec = list(prior = "pc.prec", param = c(1, .01))),
      scale.model = TRUE, replicate = TeamA, constr = FALSE) +
    f(round_away, w_roundB, copy = "round_home", replicate = TeamB)

model_rw1 <- inla(formula_rw1, data = dat_model,
                  control.predictor = list(link = 1),
                  control.compute = list(config = TRUE, cpo = TRUE,
                                         dic = TRUE, waic = TRUE),
                  family = "poisson")
```

### AR1 model

```{r}
formula_ar1 <- score ~ -1 + home.adv +
    f(round_home, w_roundA, model = "ar1",
      hyper = list(theta1 = list(prior = "pc.prec", param = c(1, .01)),
                   theta2 = list(prior = "pc.cor1", param = c(.1, .9))),
    replicate = TeamA, constr = FALSE) +
    f(round_away, w_roundB, copy = "round_home", replicate = TeamB)

model_ar1 <- inla(formula_ar1, data = dat_model,
                  control.predictor = list(link = 1),
                  control.compute = list(config = TRUE, cpo = TRUE,
                                         dic = TRUE, waic = TRUE),
                  family = "poisson")
```

### Matern model

```{r}
formula_mat <- score ~ -1 + home.adv +
    f(round_home, w_roundA,
      model = inla.spde2.pcmatern(mesh = inla.mesh.1d(1:n_rounds),
                                  alpha = 2,
                                  prior.range = c(2, .05),
                                  prior.sigma = c(1, .01)),
      replicate = TeamA, constr = FALSE) +
    f(round_away, w_roundB, copy = "round_home", replicate = TeamB)

model_mat <- inla(formula_mat, data = dat_model,
                  control.predictor = list(link = 1),
                  control.compute = list(config = TRUE, cpo = TRUE,
                                         dic = TRUE, waic = TRUE),
                  family = "poisson")
```

## Comparisons

```{r fig1, fig.width=10, fig.cap="Figure 1: A) rw1; B) ar1; C) Matern."}
top5 = c("ManchesterCity", "Liverpool", "Chelsea", "Tottenham", "Arsenal")

plot.re_rw1 <- homeperformance(teams, ploteam = top5, n_rounds,
                               mod = model_rw1)

plot.re_ar1 <- homeperformance(teams, ploteam = top5, n_rounds,
                               mod = model_ar1)

plot.re_mat <- homeperformance(teams, ploteam = top5, n_rounds,
                               mod = model_mat)

plot.re_all <- plot.re_rw1 | plot.re_ar1 | plot.re_mat
plot.re_all +
    plot_layout(guides = "collect") +
    plot_annotation(title = "Premier League Top5 performances",
                    tag_levels = "A", tag_suffix = ")")
```

```{r fig2, fig.height=3, fig.cap="Figure 2: A) rw1; B) ar1; C) Matern."}
home.adv_all <-
    plot_home.adv(model_rw1) |
    plot_home.adv(model_ar1) |
    plot_home.adv(model_mat)
home.adv_all +
    plot_annotation(title = "Marginal distribution of home advantage",
                    tag_levels = "A", tag_suffix = ")")
```

```{r}
rank_rw1 <- ranks(teams, mod = model_rw1, n_rounds, data = dat)
rank_ar1 <- ranks(teams, mod = model_ar1, n_rounds, data = dat)
rank_mat <- ranks(teams, mod = model_mat, n_rounds, data = dat)

rank_all <- cbind(rank_rw1$rank[ , "Rank by performance"],
                  rank_ar1$rank[ , "Rank by performance"],
                  rank_mat$rank)
colnames(rank_all) <- c("Rank RW1", "Rank AR1", "Rank Matern",
                        "Real rank", "Points")
print(rank_all, quote = FALSE)
```

```{r fig3, fig.width=10, fig.height=4.75, fig.cap="Figure 3: A) rw1; B) ar1; C) Matern."}
par(mar = c(7, 4, 2, 1) + .1, mfrow = c(1, 3))

plot_teams(model_rw1, teams, rank_rw1$index_order, n_rounds, type = 1)
plot_teams(model_ar1, teams, rank_ar1$index_order, n_rounds, type = 1)
plot_teams(model_mat, teams, rank_mat$index_order, n_rounds, type = 1)
```

```{r fig4, fig.width=10, fig.height=5.75, fig.cap="Figure 4: A) rw1; B) ar1; C) Matern."}
teams_rw1 <- plot_teams(model_rw1, teams, rank_rw1$index_order,
                        n_rounds, type = 2)
teams_ar1 <- plot_teams(model_ar1, teams, rank_ar1$index_order,
                        n_rounds, type = 2)
teams_mat <- plot_teams(model_mat, teams, rank_mat$index_order,
                        n_rounds, type = 2)

teams_all <- teams_rw1 | teams_ar1 | teams_mat
teams_all +
    plot_annotation(title = "Premier League team performances",
                    tag_levels = "A", tag_suffix = ")")
```

Comparison amongst the models obtained are performed using the most used
Bayesian criteria:

1. CPO (Conditional Predictive Ordinate)
2. PIT (Predictive Integral Transform)
3. DIC (Deviance Information Criteria)
4. WAIC (Watanabe Information Criteria)

The first two quantities well summarise the goodness of fit of the model
concerning the data, while the last two represent a natural counterpart
of classical comparison criteria such as AIC and BIC.

```{r}
model_obj <- list(model_rw1, model_ar1, model_mat)

inla_table <- sapply(model_obj, criteria_eval)
rownames(inla_table) <- c("CPO", "DIC", "WAIC")
colnames(inla_table) <- c("RW1", "AR1", "Matern")
inla_table
```

> The RW1 model has better measures, however, it has one less parameter.
> In the end, the models are basically equal.

```{r fig5, fig.height=3, fig.cap="Figure 5: A) rw1; B) ar1; C) Matern."}
pit_rw1 <- pit_plot(model_rw1, name = "Poisson")
pit_ar1 <- pit_plot(model_ar1, name = "Poisson")
pit_mat <- pit_plot(model_mat, name = "Poisson")

pit_all <- pit_rw1 | pit_ar1 | pit_mat
pit_all +
    plot_annotation(title = "PITs histograms",
                    tag_levels = "A", tag_suffix = ")")
```

## Simulation

```{r}
N = 1e4
tic()
model_rw1.sim <- table_producer(teams, mod = model_rw1, N, data = dat)
toc()
model_rw1.sim <- model_rw1.sim/N
model_rw1.sim <- model_rw1.sim[rank_rw1$index_true, ]
colindex <- model_rw1.sim >= .05
model_rw1.sim[] <- percent(model_rw1.sim)
model_rw1.sim <- data.frame(model_rw1.sim)
table_plot(model_rw1.sim, colindex)
```

> This table can answer one curious question.

> Is `ManchesterCity` truly better than `Liverpool` in the 2018-2019
> season, given the difference of points is just 1?

> This question is interesting because the `Manchester City` retained
> the league title on the final day of the season, finishing on 98
> points. At the same time, `Liverpool` finished runners-up with 97
> points – the highest total in English top-flight history for a
> second-place team.

> The answer to this question is `yes`, according to our model. We can
> see that the `ManchesterCity` won much more first place than
> `Liverpool` in our simulated seasons. We also notice that `Tottenham
> Hotspur` may be a better team compared to `Chelsea` based on our
> model, even though it lost by one point to `Chelsea`.
