---
title: "Modeling the cumulative incidence function of clustered competing
        risks data: \\newline
        a multinomial GLMM approach"
subtitle: master thesis defense
author: Henrique Laureano ([.github.io](https://henriquelaureano.github.io))
date: \today
institute: LEG @ UFPR
classoption: [aspectratio=169]
bibliography: ../thesis/references.bib
link-citations: yes
output:
  beamer_presentation:
    includes:
      in_header: beamerheader.txt
---

# Data

### Clustered competing risk data

Key terms:

1. \textcolor{AlertOrange}{Clustered}:
   groups with a dependence structure (e.g. families);

2. Causes \textcolor{AlertOrange}{competing} by *something*.

\bigskip
Something?

\begin{columns}
 \column{5cm}
  \begin{itemize}
  \item \textbf{Failure} of an industrial or electronic component;
 \end{itemize}
 \column{6.25cm}
  \begin{itemize}
   \item \textbf{Occurence} or \textbf{cure} of a disease or some
         biological process;
   \end{itemize}
 \column{4cm}
  \begin{itemize}
   \item \textbf{Progress} of a patient clinic state.
  \end{itemize}
\end{columns}

\bigskip
Independent of the application, always the same framework

\begin{tabular}{lc*{3}{>{\columncolor[gray]{0.8}}c}cr}
 Cluster & ID & Cause 1 & Cause 2 & Censorship & Time & Feature\\
 \hline
 1     & 1  & Yes     & No       & No        & 10   & A\\
 1     & 2  & No      & No       & Yes       &  8   & A\\
 2     & 1  & No      & No       & Yes       &  7   & B\\
 2     & 2  & No      & Yes      & No        &  5   & A\\
\end{tabular}

### Big picture: Failure time data

\bigskip
\begin{block}{}
\begin{center}
 \scalebox{0.9}{
  \begin{tikzpicture}
   \begin{pgfonlayer}{nodelayer}
    \node [style=circle] (2) at (-17.5, 5.5) {0};
    \node [style=circle] (3) at (-13.5, 5.5) {1};
    \node [style=circle] (7) at (-7.5, 5.5) {1};
    \node [style=circle] (9) at (-7.5, 4.5) {2};
    \node [style=none] (11) at (-7.5, 3.5) {$\vdots$};
    \node [style=circle] (12) at (-7.5, 2.5) {$m$};
    \node [style=circle] (14) at (-12.25, 4) {0};
    \node [style=circle] (22) at (-6, 3.5) {0};
    \node [style=circle] (23) at (-2, 3.5) {2};
    \node [style=circle] (25) at (-4, 5.5) {1};
    \node [style=none] (27) at (-17, 5.5) {};
    \node [style=none] (28) at (-14, 5.5) {};
    \node [style=none] (29) at (-11.75, 4.25) {};
    \node [style=none] (30) at (-8, 5.5) {};
    \node [style=none] (31) at (-8, 4.5) {};
    \node [style=none] (32) at (-8, 2.5) {};
    \node [style=none] (33) at (-5.5, 3.5) {};
    \node [style=none] (34) at (-4.5, 5.25) {};
    \node [style=none] (35) at (-2.5, 3.5) {};
    \node [style=none] (36) at (-11.75, 4) {};
    \node [style=none] (37) at (-11.75, 3.75) {};
    \node [style=none] (38) at (-5.5, 3.75) {};
    \node [style=none] (39) at (-2.5, 3.75) {};
    \node [style=none] (40) at (-3.5, 5.25) {};
    \node [style=none] (44) at (-15.5, 6.75) {{\large
                                               Failure time process}};
    \node [style=none] (45) at (-10, 6.75) {{\large
                                             Competing risk process}};
    \node [style=none] (47) at (-4, 6.75) {{\large Multistate process}};
   \end{pgfonlayer}
   \begin{pgfonlayer}{edgelayer}
	\draw [style=1side] (27.center) to (28.center);
	\draw [style=1side] (37.center) to (32.center);
	\draw [style=1side] (40.center) to (39.center);
	\draw [style=1side] (33.center) to (35.center);
	\draw [style=2side] (38.center) to (34.center);
	\draw [style=1side] (36.center) to (31.center);
	\draw [style=1side] (29.center) to (30.center);
   \end{pgfonlayer}
  \end{tikzpicture}
 }
\end{center}

\bigskip
\begin{columns}
 \column{4cm}
  \textit{Same methodologies, different names.}
  \column{9cm}
   \begin{description}
    \item[Survival analysis] Biomedical studies;
    \item[Reliability analysis] Industrial life testing.
   \end{description}
\end{columns}
\end{block}

\includegraphics[height=0.5cm]{logo/book.png}
A comprehensive reference is @kalb&prentice's book.

# Model

### Modeling clustered competing risks data

\bigskip
\begin{columns}
 \column{\linewidth/3}
  \centering
   \includegraphics[height=3.6cm]{logo/what.png}\\
   \bigskip
   \textcolor{UniBlue}{\textbf{What?}}
 \column{\linewidth/3}
  \centering
   \includegraphics[height=3.6cm]{logo/why.jpg}\\
   \bigskip
   \textcolor{UniBlue}{\textbf{Why?}}
 \column{\linewidth/3}
  \centering
   \includegraphics[height=3.6cm]{logo/roll_safe.png}\\
   \bigskip
   \textcolor{UniBlue}{\textbf{How?}}
\end{columns}

### Failure time data \(\rightarrow\) Survival models

First of all, we have to choose which \textcolor{UniBlue}{scale} we
model the \textbf{survival experience}.

1. Usually, is in the
   \begin{equation}
    \text{\textcolor{UniBlue}{hazard} (failure rate)
          \textcolor{UniBlue}{scale}}: \quad
    \lambda(t \mid \text{features}) =
    \lambda_{0}(t) \times c(\text{features}).\label{fig:hazard}
   \end{equation}

We have a Equation \ref{fig:hazard} for each competing cause.

The cluster dependence is something actually not measured\dots
\begin{block}{}
 \begin{center}
  Not measured dependence~\(\rightarrow\)
  \textbf{random/latent effects}~\(\rightarrow\)
  Frailty models.
 \end{center}
\end{block}

Full likelihood analysis with frailty models for competing risks data is
generally complicated, when not impracticable.

2. \textit{Not} usually, the \textcolor{UniBlue}{probability scale}.

### Probability scale \(\rightarrow\) Cause-specific CIF

Besides the within-cluster dependence, there is an often interest\newline
in describing the time at event onset, directly described by\newline
the cause-specific

```{r cif, echo=FALSE,fig.height=2,fig.width=4}
pacman::p_load(tidyverse, ## you need the pacman's package
               ggplot2)
time  <- seq(30, 79.5, by=0.5)
delta <- 80
u     <- c(0, 0)
eta   <- c(0, 0)
beta  <- c(-2, -1.5)
gamma <- c(1.2, 1)
w     <- c(3, 5)
risk1 <- exp(beta[1] + u[1])
risk2 <- exp(beta[2] + u[2])
level <- 1 +  risk1 + risk2
risklevel1 <- risk1/level
risklevel2 <- risk2/level
traj1 <- pnorm(w[1] * atanh(2 * time/delta - 1) - gamma[1] - eta[1])
traj2 <- pnorm(w[2] * atanh(2 * time/delta - 1) - gamma[2] - eta[2])
cif1 <- risklevel1 * traj1
cif2 <- risklevel2 * traj2
df <- as.data.frame(cbind(time, cif1, cif2))
df <- df%>%
    pivot_longer(cols=cif1:cif2, names_to="cif", values_to="value")
ggplot(df, aes(time, value, group=cif, color=cif))+
    geom_line(size=1.5)+
    labs(x="Time", y="Incidence",
         title='Cumulative Incidence Function (CIF)')+
    scale_color_manual(values=c("#0096c8", "#262626"),
                       labels=c("Cause 1", "Cause 2"))+
    theme(legend.title=element_blank(),
          legend.text=element_text(size=9),
          axis.text.x=element_text(size=8),
          axis.text.y=element_text(size=8),
          plot.title=element_text(
              size=10,
              margin=unit(c(t=0, r=0, b=2, l=0), 'mm')), 
          axis.title.x=element_text(
              size=9,
              margin=unit(c(t=2, r=0, b=0, l=0), "mm")),
          axis.title.y=element_text(
              size=9,
              margin=unit(c(t=0, r=2, b=0, l=0), "mm")))
```

\begin{block}{}
\[
 \text{i.e.},\quad
 \text{CIF} =
 \mathbb{P}[~\text{failure time} \leq t,
            ~\text{a given cause} \mid \text{features \& latent effects}
            ~].
\]
\end{block}

### Formally,

for a cause-specific of failure \(k\),\newline
the cumulative incidence function (CIF) is defined as
\begin{block}{}
\vspace{-0.5cm}
\begin{align*}
 F_{k}(t \mid \bm{x})
 &= \mathbb{P}[T \leq t,~K = k \mid \bm{x}]\\
 &= \int_{0}^{t} f_{k}(z \mid \bm{x})~\text{d}z\quad
 (f_{k}(t \mid \bm{x})
 ~\text{is the (sub)density for the time to a type}~k~\text{failure})\\
 &= \int_{0}^{t}
 \underbrace{\lambda_{k}(z \mid \bm{x})}_{\substack{
                                           \text{cause-specific}\\
                                           \text{hazard function}}}
 \underbrace{S(z \mid \bm{x})}_{\substack{\text{overall}\\
                                          \text{survival}\\
                                          \text{function}}}
 \text{d}z, \quad t > 0, \quad k = 1,~\dots,~K.
\end{align*}
\end{block}

\includegraphics[height=0.5cm]{logo/book.png}
Again, a comprehensive reference is @kalb&prentice's book.

\includegraphics[height=0.5cm]{logo/lightbulb.jpg}
\textit{Here}, we use the same CIF specification of @SCHEIKE.

### @SCHEIKE's CIF specification

For two competing causes of failure,\newline
the cause-specific CIFs are specified in the following manner
\begin{block}{}
\vspace{-0.5cm}
\begin{equation}
 F_{k} (t \mid \bm{x},~u_{1},~u_{2},~\eta_{k}) =
 \underbrace{\pi_{k}(\bm{x},~u_{1},~u_{2})}_{
 \substack{\text{cluster-specific}\\\text{risk level}}}\times
  \underbrace{\Phi[w_{k} g(t) - \bm{x}\bm{\gamma}_{k} - \eta_{k}]}_{
   \substack{\text{cluster-specific}\\\text{failure time trajectory}}
  }, \quad t > 0, \quad k = 1,~2,
  \label{eq:cif}
\end{equation}
\end{block}

with

1. \(\pi_{k}(\bm{x}, \bm{u}) =
     \exp\{\bm{x}\bm{\beta}_{k} + u_{k}\} \Big/
     \left(1 + \sum_{m=1}^{K-1} \exp\{\bm{x}\bm{\beta}_{m} + u_{m}\}
     \right),
     \quad k = 1,~2, \quad K = 3;\)

2. \(\Phi(\cdot)\) is the cumulative distribution function of a standard
   Gaussian distribution;

3. \(g(t) = \text{arctanh}(2t/\delta - 1),
     \quad t\in(0,~\delta), \quad g(t)\in(-\infty,~\infty).\)

\includegraphics[height=0.5cm]{logo/book.png}
In @SCHEIKE, this CIF specification is modeled under a *complicated*
pairwise composite likelihood approach [@lindsay88; @varin11].

### Our contribution: a full likelihood analysis

\vspace{-0.3cm}
For two competing causes of failure, a subject \(i\), in the cluster
\(j\), in time \(t\), we have
\begin{block}{}
\vspace{-0.5cm}
\begin{equation}
 \begin{aligned}
  y_{ijt} \mid
  \underbrace{\{u_{1j}, u_{2j}, \eta_{1j}, \eta_{2j}\}}_{
  \substack{\text{\color{UniBlue}{latent effects}}}
  } & \sim
  \text{Multinomial}(p_{1ijt}, p_{2ijt}, p_{3ijt})\\
  \begin{bmatrix} u_{1}\\ u_{2}\\ \eta_{1}\\ \eta_{2} \end{bmatrix}
  & \sim
  \parbox{2cm}{\centering Multivariate Normal}
  \left(
   \begin{bmatrix} 0\\ 0\\ 0\\ 0 \end{bmatrix},
   \begin{bmatrix}
    \sigma_{u_{1}}^{2}&
    \text{cov}(u_{1}, u_{2})&
    \text{cov}(u_{1}, \eta_{1})&\text{cov}(u_{1}, \eta_{2})\\
    &\sigma_{u_{2}}^{2}&
    \text{cov}(u_{2}, \eta_{1})&\text{cov}(u_{2}, \eta_{2})\\
    &&\sigma_{\eta_{1}}^{2}&\text{cov}(\eta_{1}, \eta_{2})\\
    &&&\sigma_{\eta_{2}}^{2}
   \end{bmatrix}
  \right)\\  
  p_{ {\color{UniBlue}{k}} ijt} &=
  \frac{\partial}{\partial t}
  F_{\color{UniBlue}{k}}
  (t \mid \bm{x}, \bm{u}, \eta_{\color{UniBlue}{k}})\label{eq:model}\\
  &= \frac{\exp\{\bm{x}_{ {\color{UniBlue}{k}} ij}
                 \bm{\beta}_{\color{UniBlue}{k}} +
                 u_{ {\color{UniBlue}{k}} j}\}}{
  1 + \sum_{ {\color{UniBlue}{m}}=1}^{K-1}
      \exp\{\bm{x}_{ {\color{UniBlue}{m}} ij}
            \bm{\beta}_{\color{UniBlue}{m}} +
            u_{ {\color{UniBlue}{m}} j}\}}\\
  &\times w_{\color{UniBlue}{k}}\frac{\delta}{2\delta t - 2t^{2}}~
  \phi\left(
   w_{\color{UniBlue}{k}}
   \text{arctanh}\left(\frac{t-\delta/2}{\delta/2}\right)
   - \bm{x}_{ {\color{UniBlue}{k}} ij}
     \bm{\gamma}_{\color{UniBlue}{k}} -
     \eta_{ {\color{UniBlue}{k}} j}
  \right), \quad {\color{UniBlue}{k}} = 1,~2.
 \end{aligned}
\end{equation}
\end{block}

### Simulating from the model

\vspace{0.2cm}
```{r simu,echo=FALSE,fig.height=4.9}
pacman::p_load(tidyverse, ## you need the pacman's package
               ggplot2,
               mvtnorm,
               mc2d, 
               patchwork)
time  <- seq(30, 79.5, by=0.5)
delta <- 80
u     <- c(0, 0)
eta   <- c(0, 0)
beta  <- c(-2, -1.5)
gamma <- c(1.2, 1)
w     <- c(3, 5)
risk1 <- exp(beta[1] + u[1])
risk2 <- exp(beta[2] + u[2])
level <- 1 +  risk1 + risk2
risklevel1 <- risk1/level
risklevel2 <- risk2/level
traj1 <- pnorm(w[1] * atanh(2 * time/delta - 1) - gamma[1] - eta[1])
traj2 <- pnorm(w[2] * atanh(2 * time/delta - 1) - gamma[2] - eta[2])
cif1 <- risklevel1 * traj1
cif2 <- risklevel2 * traj2
df <- as.data.frame(cbind(time, cif1, cif2))
df <- df%>%
    pivot_longer(cols=cif1:cif2, names_to="cif", values_to="value")
x1 <- w[1] * atanh(2 * time/delta - 1) - gamma[1] - eta[1]
x2 <- w[2] * atanh(2 * time/delta - 1) - gamma[2] - eta[2]
dtraj1 <- w[1] * delta/(2 * time * (delta - time)) * dnorm(x1)
dtraj2 <- w[2] * delta/(2 * time * (delta - time)) * dnorm(x2)
dcif1 <- risklevel1 * dtraj1
dcif2 <- risklevel2 * dtraj2
ddf <- as.data.frame(cbind(time, dcif1, dcif2))
ddf <- ddf%>%
    pivot_longer(cols=dcif1:dcif2, names_to="dcif", values_to="value")
g1 <-
    ggplot(df, aes(time, value, group=cif, color=cif))+
    geom_line(size=1.5)+
    labs(x="Time", y=NULL)+
    scale_color_manual(values=c("#0096c8", "#262626"),
                       labels=c("CIF 1", "CIF 2"))+
    theme(legend.title=element_blank(),
          legend.text=element_text(size=13),
          axis.text.x=element_text(size=11),
          axis.text.y=element_text(size=11),
          axis.title.x=element_text(
              size=12,
              margin=unit(c(t=2, r=0, b=0, l=0), "mm")))
g2 <-
    ggplot(ddf, aes(time, value, group=dcif, color=dcif))+
    geom_line(size=1.5)+
    labs(x="Time", y=NULL)+
    scale_color_manual(values=c("#0096c8", "#262626"),
                       labels=c("dCIF 1", "dCIF 2"))+
    theme(legend.title=element_blank(),
          legend.text=element_text(size=13),
          axis.text.x=element_text(size=11),
          axis.text.y=element_text(size=11),
          axis.title.x=element_text(
              size=12,
              margin=unit(c(t=2, r=0, b=0, l=0), "mm")))
J    <- 10e3
cs   <- 2
time <- runif(n=cs*J, min=30, max=79.9)
Z    <- Matrix::bdiag(replicate(J, rep(1, cs), simplify=FALSE))
S    <- matrix(c( 1.0,  0.4, -0.1,  0.4,
                  0.4,  1.0,  0.4, -0.1,
                 -0.1,  0.4,  1.0,  0.4,
                  0.4, -0.1,  0.4,  1.0), 4)
out <- tibble::tibble(i=rep(seq(cs), times=J),
                      j=rep(seq(J),  each=cs),
                      time=time,
                      p1=NA, p2=NA, p3=NA)
K     <- dim(S)[1]/2 + 1
ladim <- 2*(K-1) ## latent effects dimension
set.seed(11)
U     <- mvtnorm::rmvnorm(J, mean=rep(0, ladim), sigma=S)
ZU    <- Z%*%U
risk1 <- exp(beta[1] + ZU[, 1])
risk2 <- exp(beta[2] + ZU[, 2])
level <- 1 + risk1 + risk2
gt    <- atanh(2*time/delta - 1)
dgt   <- delta/(2*time*(delta - time))
x1    <- w[1]*gt - gamma[1] - ZU[, 3]
x2    <- w[2]*gt - gamma[2] - ZU[, 4]
out$p1 <- risk1/level*w[1]*dgt*dnorm(x1)
out$p2 <- risk2/level*w[2]*dgt*dnorm(x2)
out    <- out%>%dplyr::mutate(p3=1-p1-p2)
set.seed(31)
y      <- mc2d::rmultinomial(cs*J, 1, prob=out%>%select(p1:p3))
out    <- out%>%
    dplyr::bind_cols(tibble::as_tibble(y))%>%
    dplyr::rename(y1=V1, y2=V2, y3=V3)
datp   <- out%>%
    pivot_longer(cols=p1:p3, names_to='p_label', values_to='p_value')%>%
    mutate(p_label=forcats::fct_recode(p_label,
                                       'Cause 1'='p1',
                                       'Cause 2'='p2',
                                       'Censorship'='p3'))
daty   <- out%>%
    pivot_longer(cols=y1:y3, names_to='y_label', values_to='y_value')%>%
    filter(y_value==1)
g3 <- 
    ggplot(datp, aes(x=p_value))+
    geom_histogram(binwidth=0.0025, color='white', fill="#0096c8")+
    facet_wrap(~p_label, scales='free')+
    labs(x='Probability', caption='bindwidth=0.0025')+
    geom_label(data=data.frame(
                   label=paste0(
                       "Turns out in\n",
                       round(prop.table(table(daty$y_label)), 3),
                       "%'s of data"),
                   p_label=levels(datp$p_label)),
               mapping=aes(x=c(0.025, 0.035, 0.962),
                           y=Inf, label=label),
               vjust=1.5)+
    ## theme_minimal()+
    theme(strip.background=element_rect(colour='black', fill='white'),
          strip.text.x=element_text(size=13), 
          axis.text.x=element_text(size=11),
          axis.text.y=element_blank(),
          axis.title.y=element_blank(),
          axis.title.x=element_text(
              size=12, 
              margin=unit(c(t=3, r=0, b=0, l=0), 'mm')),
          plot.caption=element_text(size=11))
(g1+g2)/g3
```

### Marginal likelihood function for two competing causes

\vspace{-0.75cm}
\begin{align}
 L(\bm{\theta}; y)
 &= \prod_{j=1}^{J}~\int_{\Re^{4}}
    \pi(y_{j} \mid \bm{r}_{j})\times\pi(\bm{r}_{j})~\text{d}\bm{r}_{j}
    \nonumber\\
 &= \prod_{j=1}^{J}~\int_{\Re^{4}}
    \Bigg\{
    \underbrace{\prod_{i=1}^{n_{j}}~\prod_{t=1}^{n_{ij}}
    \Bigg(
    \frac{(\sum_{k=1}^{K}y_{kijt})!}{y_{1ijt}!~y_{2ijt}!~y_{3ijt}!}~
    \prod_{k=1}^{K} p_{kijt}^{y_{kijt}}
    \Bigg)}_{\substack{\text{\color{UniBlue}{fixed effect component}}}}
  \Bigg\}\times\nonumber\\
 &\hspace{2cm}\underbrace{
   (2\pi)^{-2} |\Sigma|^{-1/2} \exp
   \left\{-\frac{1}{2}\bm{r}_{j}^{\top} \Sigma^{-1} \bm{r}_{j}\right\}
   }_{\substack{\text{\color{UniBlue}{latent effect component}}}}
   \text{d}\bm{r}_{j}\nonumber\\
 &= \prod_{j=1}^{J}~\int_{\Re^{4}}
    \Bigg\{
    \underbrace{\prod_{i=1}^{n_{j}}~\prod_{t=1}^{n_{ij}}
    \prod_{k=1}^{K} p_{kijt}^{y_{kijt}}
    }_{\substack{\color{UniBlue}{\text{fixed effect}}}}
   \Bigg\}\underbrace{
   (2\pi)^{-2} |\Sigma|^{-1/2} \exp
   \left\{-\frac{1}{2}\bm{r}_{j}^{\top} \Sigma^{-1} \bm{r}_{j}\right\}
   }_{\substack{\color{UniBlue}{\text{latent effect component}}}}
   \text{d}\bm{r}_{j}\label{eq:loglik},
\end{align}
with \(p_{kijt}\) from Equation \ref{eq:model} and where \(\bm{\theta} =
[\bm{\beta}~\bm{\gamma}~\bm{w}~\bm{\sigma^{2}}~ \bm{\rho}]^{\top}\) is
the parameters vector.

# TMB: Template Model Builder

### TMB: Automatic Differentiation and Laplace Approximation

\includegraphics[height=0.5cm]{logo/book.png}
@TMB.

\bigskip
> An R [@R21] package for the quickly implementation of complex random
> effect models through simple C++ templates.

\begin{block}{Workflow}
 \begin{enumerate}
  \item Write your objective function in a \texttt{.cpp} through a
        \texttt{\#include <TMB.hpp>};

  \vspace{0.15cm}
  \item Compile and load it in R via \texttt{TMB::compile()} and
        \texttt{base::dyn.load(TMB::dynlib())};

  \vspace{0.15cm}
  \item Compute your objective function derivatives with
        \texttt{obj <- TMB::MakeADFun()};

  \vspace{0.15cm}
  \item Perform the model fitting,
        \texttt{opt <- base::nlminb(obj\$par, obj\$fn, obj\$gr)};

  \vspace{0.15cm}
  \item Compute the parameters standard deviations,
        \texttt{TMB::sdreport(obj)}.
 \end{enumerate}
\end{block}

\includegraphics[height=0.5cm]{logo/book.png}
For details about TMB, AD, and Laplace approximation: @laurence.

# Simulation study

### Simulation study model designs

\begin{columns}
 \column{5.75cm}
  \begin{block}{Risk model}
   Latent effects only on the risk level i.e.,
   \[
    \Sigma = \begin{bmatrix}
              \sigma_{u_{1}}^{2} & \text{cov}_{u_{1}, u_{2}}\\
                                 & \sigma_{u_{2}}^{2}
             \end{bmatrix}.
   \]
  \end{block}
 \column{5.75cm}
  \begin{block}{Time model}
   Latent effects only on the failure time trajectory level i.e.,
   \[
    \Sigma = \begin{bmatrix}
              \sigma_{\eta_{1}}^{2} & \text{cov}_{\eta_{1}, \eta_{2}}\\
                                    & \sigma_{\eta_{2}}^{2}
              \end{bmatrix}.
   \]
  \end{block}
\end{columns}

\begin{columns}
 \column{7cm}
  \begin{block}{Block-diag model}
   Latent effects on the risk and time levels without cross-correlations
   i.e.,
   \[
    \Sigma = \begin{bmatrix}
              \sigma_{u_{1}}^{2} & \text{cov}_{u_{1}, u_{2}} & 0 & 0\\
              & \sigma_{u_{2}}^{2} & 0 & 0\\
              & & \sigma_{\eta_{1}}^{2} & \text{cov}_{\eta_{1}, \eta_{2}}\\
              & & & \sigma_{\eta_{2}}^{2}
             \end{bmatrix}.
   \]
  \end{block}
 \column{7cm}
  \begin{block}{Complete model}
   A \textit{complete} latent effects structure\newline
   i.e.,
   \[
    \Sigma = \begin{bmatrix}
              \sigma_{u_{1}}^{2} &
              \text{cov}_{u_{1}, u_{2}} &
              \text{cov}_{u_{1}, \eta_{1}} & \text{cov}_{u_{1}, \eta_{2}}\\
              & \sigma_{u_{2}}^{2} &
              \text{cov}_{u_{2}, \eta_{1}} & \text{cov}_{u_{2}, \eta_{2}}\\
              & & \sigma_{\eta_{1}}^{2} & \text{cov}_{\eta_{1}, \eta_{2}}\\
              & & & \sigma_{\eta_{2}}^{2}
              \end{bmatrix}.
   \]
  \end{block}
\end{columns}

### Simulation study setup

# Conclusion

### Thanks for watching and have a great day

\vspace{0.5cm}
Special thanks to

\bigskip
\begin{columns}
 \column{1.5cm}
  \centering\includegraphics{logo/ppgmne-logo.png}
 \column{13.75cm}
  {\large \textbf{PPGMNE}}\newline
  Programa de Pós-Graduação em\newline
  Métodos Numéricos em Engenharia
\end{columns}

\bigskip
\begin{columns}
 \column{2cm}
  \includegraphics[height=2cm]{logo/capes.jpg}
 \column{3cm}
  \includegraphics[height=2cm]{logo/ufpr.png}
 \column{4.5cm}
  Joint work with
  \vspace{0.2cm}\newline
  Wagner H. Bonat\newline
  \url{http://leg.ufpr.br/~wagner}
 \column{4.75cm}
  \textcolor{white}{Joint work with}
  \vspace{0.2cm}\newline
  Paulo Justiniano Ribeiro Jr.\newline
  \url{http://leg.ufpr.br/~paulojus}
\end{columns}

\vspace{1.5cm}
\begin{columns}
 \column{10cm}
 \column{5cm}
  \includegraphics[height=0.3cm,angle=90]{../../laurence.jpg}
  \hspace{0.05cm}
  \href{https://henriquelaureano.github.io}{henriquelaureano.github.io}
\end{columns}

# References

### References

\small
