---
title: "An analysis of the union wages data: \\newline GLM's, GAM's and JAGS"
short-title: "Project presentation (the project isn't done yet!)"
author: "Henrique Laureano \\newline \\texttt{mynameislaure.github.io}"
short-author: "Henrique Laureano"
# email: "henrique.laureano\\@kaust.edu.sa"
# date: ""
# short-date: ""                              # `r format(Sys.Date(), "%m/%d/%Y")`
institute: "STAT 260: Nonparametric Statistics"
short-institute: "STAT 260, KAUST"
# department: ""
section-titles: true                    # provides slide headings
safe-columns: true                      # enables special latex macros for columns
fontsize: 10pt
output:
  beamer_presentation:
    template: template.tex
---

# Data

***

Turning the dataset into a Kandinsky* painting

```{r, echo=FALSE}
data(trade.union, package = "SemiPar")

kandinsky::kandinsky(trade.union)
```

\hfill (github.com/gsimchoni/kandinsky)

***

\* Kandinsky

\begincols
 \column{.45\linewidth}
  \begin{figure}
   \centering
   \includegraphics[width=\textwidth]{google_kandinsky.png}
  \end{figure}
 \column{.55\linewidth}
  \begin{figure}
   \centering
   \includegraphics[width=\textwidth]{kandinsky_flavor.png}
  \end{figure}
\endcols

\hfill {\footnotesize (screenshots from Google)}

***

\begin{block}{Trade union data}

Data on 534 U.S. workers with eleven variables (\texttt{SemiPar::trade.union}).

\end{block}

\pause

Variables:

\pause

\begincols

\column{.4\linewidth}

- \color{blue}{\texttt{union.member}} \color{black} \newline {\small (yes or no)}

- \texttt{years.educ}

- \texttt{years.experience}

- \texttt{wage} \newline {\small (dollars per hour)}

- \texttt{age}

- \texttt{female} \newline {\small (yes or no)}

\column{.6\linewidth}

- \texttt{south} \newline {\small (living or not in southern region of U.S.)}

- \texttt{race} \newline {\small (black, hispanic or white)}

- \texttt{occupation} \newline {\small (six categories)}

- \texttt{sector} \newline {\small (three categories)}

- \texttt{married} \newline {\small (yes or no)}

\endcols

***

Quantitative variables:

```{r, echo=FALSE}
pairs(trade.union[ , c(1, 4, 6:7)]
      , upper.panel = function(x, y, ...) {
        usr <- par("usr") ; on.exit(par(usr)) ; par(usr = c(0, 1, 0, 1))
        r <- abs(cor(x, y))
        txt <- format(c(r, 0.123456789), digits = 2)[1]
        text(.5, .5, txt, cex = 2)
      }
      , pch = 19, gap = .25, xaxt = "n", yaxt = "n"
      , col = c("#0080FF", "#F3953E")[trade.union$union.member + 1])
```

\hfill {\small (colors by \texttt{union.member} status)}

# GLM

## Fitting Generalized Linear Models

\pause

- Let \(p_{i}\) be the probability of trade union membership;

- Using a logistic regression model

\vspace{-.55cm}
\begin{align*}
 {\rm logit}(p_{i}) = \beta_{0} + \beta_{1} \texttt{years.educ}_{i} + ... +
                                    \beta_{10} \texttt{married}_{i}, \\
 \texttt{union.member}_{i} \sim {\rm Bernoulli}(p_{i}), \quad i = 1, \dots 534.
\end{align*}

\hfill {\small (b/c we have 10 variables, as previously shown)}

\pause
\small
```{r, eval=FALSE}
formula <- union.member ~
  years.educ + years.experience + wage + age + female + south +
  as.factor(race) + as.factor(occupation) + sector + married

union.glm <- glm(formula, family = binomial, trade.union)
```
\normalsize	

Using the AIC as criterion we have ...

***

```{r, include=FALSE}
formula <- union.member ~
  years.educ + years.experience + wage + age + female + south +
  as.factor(race) + as.factor(occupation) + sector + married

union.glm <- glm(formula, family = binomial, trade.union)

union.glm <- MASS::stepAIC(union.glm)
```

\small
```{r, comment=NA}
union.glm$formula
```
\normalsize

\hfill we *finish* with seven variables, two quantitatives.

\pause
... and the residues? \pause ... and the goodness-of-fit? \pause

\small
```{r, message=FALSE}
pearson <- residuals(union.glm, type = "pearson")
devi <- residuals(union.glm, type = "deviance")
rocurve <- pROC::roc(trade.union$union.member, fitted(union.glm))
```

```{r, echo=FALSE, message=FALSE, fig.height=3.25}
par(mfrow = c(1, 3), mar = c(2, 3, 3, 1))

plot(pearson
     , axes = FALSE, xlab = NA, ylab = NA, main = "Pearson residuals", pch = 20)
abline(h = 0, col = 2, lwd = 1.5) ; Axis(c(-2, 8), side = 2, las = 1)

plot(devi
     , axes = FALSE, xlab = NA, ylab = NA, main = "Deviance residuals", pch = 20)
abline(h = 0, col = 2, lwd = 1.5) ; Axis(c(-2, 3), side = 2, las = 1)

library(pROC)
plot.roc(rocurve, print.auc = TRUE, print.thres = TRUE
         , las = 1, print.thres.cex = 1, print.auc.cex = 1)
```

## Coefficients

\pause
\small
```{r, comment=NA}
round( summary(union.glm)$coeff, 5)
```

```{r}
                    # null.deviance: 503.0841, deviance: 426.8709
```

## Effects

\pause
```{r, include=FALSE}
df <- trade.union

df[ , c(2:3, 5, 8:11)] <- lapply(df[ , c(2:3, 5, 8:11)], factor)

union.glm <- glm(union.member ~ ., family = binomial, df)

union.glm <- MASS::stepAIC(union.glm)

eff <- effects::allEffects(union.glm)
plot(eff, ylab = expression(p[i]), main = NULL)
```

\vspace{-.35cm}
\begin{figure}
 \centering
 \includegraphics[height=7.75cm]{effects.png}
\end{figure}

# GAM

***

## Fitting Generalized Additive Models

\pause

Logistic regression model

\vspace{-.425cm}
\begin{align*}
 {\rm logit}(p_{i}) = \beta_{0}
                    & + f_{1}(\texttt{years.educ}_{i}) + ... +
                        f_{4}(\texttt{age}_{i}) \\
                    & + \beta_{1} \texttt{female}_{i} + ... +
                        \beta_{6} \texttt{married}_{i}, \\
 & \hspace{-2cm}
 \texttt{union.member}_{i} \sim {\rm Bernoulli}(p_{i}), \quad i = 1, \dots 534.
\end{align*}

\hfill (4 quantitative variables, thus 4 smooth functions/splines, and 6,
        remaining, qualitative variables.)

\pause
\small
```{r, message=FALSE}
formula <- union.member ~
  s(years.educ) + s(years.experience, k = 20) + s(wage, k = 20) +
  s(age, k = 20) + female + south + race + occupation + sector +
  married

union.gam <- mgcv::gam(formula, family = binomial, trade.union)
```
\normalsize	

\pause
Selecting a model looking to trade off between degree of freedom and RSS ...

***

Doing variable selection in qualitative features and looking to the qualitative
ones ...

\pause
```{r, include=FALSE}
formula <- union.member ~
  s(years.educ) + s(years.experience, k = 20) + s(wage, k = 20) + s(age, k = 20) +
  female + south + as.factor(race) + as.factor(occupation)

library(mgcv)
union.gam <- gam(formula, family = binomial, trade.union)

par(mfrow = c(2, 2), mar = c(4.25, 4, 1.25, 1) + .1)

plot(union.gam, select = 1, lwd = 2, ylim = c(-4.5e5, 4.5e5)
     , xlab = "Years of education", ylab = "Spline")
plot(union.gam, select = 2, lwd = 2, ylim = c(-1.5e6, 1.5e6)
     , xlab = "Years of experience", ylab = "Spline")
plot(union.gam, select = 3, lwd = 2, ylim = c(-10, 10)
     , xlab = "Wage", ylab = "Spline")
plot(union.gam, select = 4, lwd = 2, ylim = c(-1.1e6, 1.1e6)
     , xlab = "Age", ylab = "Spline")
```

\vspace{-.25cm}
\begin{figure}
 \centering
 \includegraphics[height=7.5cm]{bad_first.png}
\end{figure}

***

\small
```{r, comment=NA}
round( anova(union.gam)$s.table, 5)
```
\normalsize

\pause
```{r, include=FALSE}
formula <- union.member ~
  s(wage, k = 20) + female + south + as.factor(race) + as.factor(occupation)

union.gam <- gam(formula, family = binomial, trade.union)

par(mar = c(4.25, 4, 1.25, 1) + .1)

plot(union.gam, lwd = 2, ylim = c(-10, 10), xlab = "Wage", ylab = "Spline")
```

\vspace{-.3cm}
\begincols
 \column{.4\linewidth}
  Doing variable selection in the qualitative's...
 \column{.6\linewidth}
  \begin{figure}
   \centering
   \includegraphics[height=5.75cm]{baaaad.png}
  \end{figure}
\endcols

## Residues

\small
```{r, comment=NA}
union.gam$formula
```

```{r}
pearson <- residuals(union.gam, type = "pearson")
devi <- residuals(union.gam, type = "deviance")
rocurve <- roc(trade.union$union.member, fitted(union.gam))
```

```{r, echo=FALSE, fig.height=3.25}
par(mfrow = c(1, 3), mar = c(2, 3, 3, 1))

plot(pearson
     , axes = FALSE, xlab = NA, ylab = NA, main = "Pearson residuals", pch = 20)
abline(h = 0, col = 2, lwd = 1.5) ; Axis(c(-2, 8), side = 2, las = 1)

plot(devi
     , axes = FALSE, xlab = NA, ylab = NA, main = "Deviance residuals", pch = 20)
abline(h = 0, col = 2, lwd = 1.5) ; Axis(c(-2, 3), side = 2, las = 1)

plot.roc(rocurve, print.auc = TRUE, print.thres = TRUE, las = 1
         , print.thres.cex = 1, print.auc.cex = 1)
```

## Coefficients

\small
```{r, comment=NA}
round( summary(union.gam)$p.table, 5)

summary(union.gam)$s.table
```

# JAGS

## Logistic regression model

\vspace{-.4cm}
\[
 {\rm logit}(p_{i}) = f(\texttt{wage}_{i}), \quad
 \texttt{union.member}_{i} \sim {\rm Bernoulli}(p_{i}), \quad
 i = 1, \dots 534.
\]

JAGS model specification file

\small
```{r, eval=FALSE}
model {
  eta <- X %*% b
  for (i in 1:n) { mu[i] <-  ilogit(eta[i]) } # expected response
  for (i in 1:n) { y[i] ~ dbin(mu[i], w[i])            # response
  for (i in 1:1) { b[i] ~ dnorm(0, .018) }         # tau=1/7.5**2
                                              # prior for s(wage)
  K1 <- S1[1:19, 1:19] * lambda[1]  + S1[1:19, 20:38] * lambda[2]
  b[2:20] ~ dmnorm(zero[2:20] ,K1) 
                                     # smoothing parameter priors
  for (i in 1:2) {
    lambda[i] ~ dgamma(.05, .005)
    rho[i] <- log(lambda[i])
  }
}
```

## Results

\small
Simulating from the model and addying a sample of 20 curves from the posterior.

```{r, include=FALSE, cache=TRUE, cache.path='cache_jags/'}
library(rjags)

jags.gam <- jagam(union.member ~ s(wage, k = 20), trade.union, family = binomial
                  , file = "trade_union.jags")

union.jags <- jags.model("trade_union.jags"
                         , data = jags.gam$jags.data, inits = jags.gam$jags.ini
                         , n.chains = 3)

samp <- jags.samples(union.jags, c("b", "rho", "mu"), n.iter = 1e4, thin = 10)

jam <- sim2jam(samp, jags.gam$pregam)
```

```{r, echo=FALSE}
plot(jam
     , shade = TRUE, shade.col = "orange", shift = coef(jam)[1]
     , trans = binomial()$linkinv, rug = FALSE, seWithMean = TRUE
     , xlim = c(0, 30), lwd = 4, col = "#0080ff", xlab = "Wage", ylab = "Spline")

nu <- trade.union$union.member == 0
with(trade.union, points(wage[nu], 0 * wage[nu], pch = 4))
with(trade.union, points(wage[!nu], 0 * wage[!nu] + .5, pch = 4))

ii <- 1:20 * 50; df.pred <- data.frame(wage = 0:300 / 10)
Xp <- predict(jam, type = "lpmatrix", newdata = df.pred)
for (i in ii) {
  p <- binomial()$linkinv(Xp %*% samp$b[, i, 1])
  lines(df.pred$wage, p, lty = 2)
}
```

## and is this...

\vspace{1.25cm}
\centering {\Large thank you!}
\vspace{1.25cm}

\begin{figure}
   \centering
   \includegraphics[width=.325\textwidth]{logo.pdf}
  \end{figure}

\url{henrique.laureano@kaust.edu.sa}