---
title: "PROTF"
author: "Henrique Laureano"
format: html
editor: visual
code-line-numbers: true
code-fold: true
code-block-border-left: true
date-modified: 2026/04/03
editor_options: 
  chunk_output_type: console
toc: true
---

```{r pkgs}

if (!requireNamespace("pacman", quietly = TRUE)) 
  install.packages("pacman") 

pacman::p_load(
  ggplot2, googlesheets4, 
  dplyr, lubridate, tidyr, stringr, ggpubr, segmented, broom, pROC, 
  knitr, purrr, patchwork
)
ggplot2::theme_set(
  theme_classic(base_size = 14) +
    theme(plot.title           = element_text(size = 14, face = "bold"),
          plot.subtitle        = element_text(size = 13, face = "bold"),
          axis.text.x          = element_text(face = "bold"),
          axis.text.y          = element_text(face = "bold"),
          axis.title.x         = element_text(hjust = 0),
          axis.title.y         = element_text(hjust = 0, angle = 360),
          strip.text           = element_text(size = 13, face = "bold"),
          legend.position      = "bottom", 
          legend.justification = "right",
          panel.grid.major     = element_line(color = "gray90", linewidth = 0.7))
)

```

```{r protf_data, warning=FALSE, message=FALSE}

protf_url <- 
  paste0(
    "https://docs.google.com/spreadsheets/d/",
    "1k3vlSk4e4mHMwTHZVnDpSTXyMReXLTGsgk82rHVILgQ/",
    "edit?gid=402323031#gid=402323031"
)
protf <- googlesheets4::read_sheet(protf_url) |> dplyr::distinct()

protf <- protf |> 
  dplyr::mutate(
    IDADE = lubridate::time_length(
      lubridate::interval(`6 - Data de Nascimento`, 
                          `5 - Data de Cadastro do Exame`), "years"
    )
  )
```

# 1. Análise epidemiológica

## a) Porcentagem de cada faixa etária na amostra

```{r}

protf |> dplyr::count(`3 - Insira Faixa etária do paciente`) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

## b) Setores mais representativos na amostra

```{r}

protf |> dplyr::count(`2 - Insira o Setor de Solicitação do exame`) |>
  dplyr::mutate(Percentual = n / nrow(protf)) |>
  dplyr::arrange(dplyr::desc(n)) |> print(n = 28)

```

## c) Sexo

```{r}

protf |> dplyr::count(`4 - Sexo`) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

## d) Presença/ausência de quadros infecciosos e sistema mais afetado em casos positivos para quadros infecciosos

```{r}

protf |> dplyr::count(`7 - Quadros infecciosos até o momento da coleta?`) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

Em quem tem caso positivo para quadro infeccioso

```{r}

protf |> dplyr::filter(
  `7 - Quadros infecciosos até o momento da coleta?` == "Sim"
) |>
  dplyr::select(`8 - Infecção em sistema nervoso central?`,
                `9 - Infecção de vias áreas superiores?`,
                `10 - Infecção de vias áreas inferiores?`,
                `11 - Infecção de sistema urinário?`,
                `12 - Infecção de sistema gastrointestinal?`,
                `13 - Infecção de pele?`,
                `14 - Sepse?`,
                `15 - Outros?`) |>
  tidyr::pivot_longer(`8 - Infecção em sistema nervoso central?`:`15 - Outros?`) |>
  dplyr::filter(value == "Sim") |>
  dplyr::count(name) |>
  dplyr::mutate(Percentual = n / 373) |> dplyr::arrange(dplyr::desc(n))

```

Dos 373 pacientes com quadro infeccioso, 154 apresentaram em mais de um sistema (ao menos em dois). 41.29%.

Nestes 154, estes são os sintomas que aparecem e em quantos dos 154 pacientes.

```{r}

proxy <- 
  protf |> dplyr::filter(
    `7 - Quadros infecciosos até o momento da coleta?` == "Sim"
  ) |>
  dplyr::select(`1 - Insira o Prontuário do Paciente`,
                `8 - Infecção em sistema nervoso central?`,
                `9 - Infecção de vias áreas superiores?`,
                `10 - Infecção de vias áreas inferiores?`,
                `11 - Infecção de sistema urinário?`,
                `12 - Infecção de sistema gastrointestinal?`,
                `13 - Infecção de pele?`,
                `14 - Sepse?`,
                `15 - Outros?`) |>
  tidyr::pivot_longer(!`1 - Insira o Prontuário do Paciente`) |>
  dplyr::filter(value == "Sim") |>
  dplyr::group_by(`1 - Insira o Prontuário do Paciente`) |>
  dplyr::mutate(n = dplyr::n()) |>
  dplyr::ungroup() |>
  dplyr::filter(n > 1) |>
  dplyr::arrange(dplyr::desc(n), `1 - Insira o Prontuário do Paciente`) 

# proxy |> dplyr::distinct(`1 - Insira o Prontuário do Paciente`)

(
  proxy2 <- 
    proxy |>
    dplyr::select(!c(value, n)) |> 
    mutate(name = stringr::str_remove(name, "^[0-9]+ - "),
           name = stringr::str_remove(name, "\\?"),
           name = stringr::str_remove(name, "Infecção (de|em) "),
           name = stringr::str_to_sentence(name)) |>
    dplyr::count(name) |> 
    dplyr::arrange(dplyr::desc(n)) |>
    dplyr::mutate(Percentual = n / 154) 
)

```

O máximo de infecções que um paciente apresenta são 6, apenas 1 paciente se apresenta em tal situação. Com 5 temos 3 pacientes.

```{r fig.height=10}

proxy |>
  dplyr::mutate(value = ifelse(value == "Sim", 1, 0)) |>
  tidyr::pivot_wider(names_from = name, values_from = value, values_fill = 0) |>
  dplyr::select(-n) |> 
  tidyr::pivot_longer(-`1 - Insira o Prontuário do Paciente`, 
                      names_to = "Infection_Type", 
                      values_to = "Present") |>
  mutate(Infection_Type = stringr::str_remove(Infection_Type, "^[0-9]+ - "),
         Infection_Type = stringr::str_remove(Infection_Type, "\\?"),
         Infection_Type = stringr::str_remove(Infection_Type, "Infecção (de|em) "),
         Infection_Type = stringr::str_to_sentence(Infection_Type)) |>
  # dplyr::filter(Present == 1) |>
  # dplyr::count(`1 - Insira o Prontuário do Paciente`) |> dplyr::arrange(dplyr::desc(n))
  ggplot() +
  aes(
    x = factor(Infection_Type, levels = unique(proxy2$name)), 
    y = factor(
      `1 - Insira o Prontuário do Paciente`, 
      levels = unique(proxy$`1 - Insira o Prontuário do Paciente`) |> rev()
    ), 
    fill = as.factor(Present)
  ) +
  geom_tile(color = "white") + # Adds a grid line effect
  scale_fill_manual(values = c("0" = "grey95", "1" = "steelblue"),
                    labels = c("Não", "Sim"),
                    name = "Presença: ") +
  labs(x = "Tipo de Infecção", 
       y = "Paciente",
       title = "Mapa de Calor de Infecções por Paciente",
       subtitle = "Apenas pacientes com ao menos dois tipos de infecção") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_discrete(labels = NULL)

```

## e) Porcentagem de indivíduos com comorbidade e comorbidades mais frequentes

```{r}

protf |> 
  dplyr::count(`20 - Paciente possui alguma comorbidade/diagnostico de base?`) |>
  dplyr::mutate(Percentual = n / nrow(protf))

protf |> 
  dplyr::filter(
    `20 - Paciente possui alguma comorbidade/diagnostico de base?` == "Sim"
  ) |>
  dplyr::count(`20.1 -  Qual comorbidade o paciente possui?

-Em caso negativo responder com NA`) |> 
  dplyr::mutate(Percentual = n / 632) |>
  dplyr::arrange(dplyr::desc(n)) |> 
  dplyr::filter(n > 1) |>
  print(n = 50)
  
```

Entre os 632 pacientes com comorbidade, 187 diferentes combinações de comorbidade são observadas - mas 47 aparecem em ao menos dois pacientes.

Mesma coisa mas agora reunindo todos os resultados similares em um único grupo de soma.

92 diferentes, 37 aparecendo ao menos duas vezes.

```{r}

protf |> 
  dplyr::filter(
    `20 - Paciente possui alguma comorbidade/diagnostico de base?` == "Sim"
  ) |>
  dplyr::select(Comorbidades = `20.1 -  Qual comorbidade o paciente possui?

-Em caso negativo responder com NA`) |>
  tidyr::separate_longer_delim(cols = Comorbidades, delim = ";") |>
  dplyr::mutate(
    Comorbidades = stringr::str_trim(Comorbidades),
    Comorbidades = stringi::stri_trans_general(Comorbidades, "Latin-ASCII"),
    Comorbidades = stringr::str_to_sentence(Comorbidades)
  ) |>
  dplyr::count(Comorbidades) |> 
  dplyr::arrange(dplyr::desc(n)) |> 
  # dplyr::filter(n > 1) |>
  print(n = 100)
  
```


## f) Porcentagem de pacientes que realizam hemodiálise

```{r}

protf |> 
  dplyr::count(`21 - Paciente realiza hemodiálise?`) |>
  dplyr::mutate(Percentual = n / nrow(protf))
  
```

## g) Porcentagem de pacientes que passaram por cirurgia nos 30 dias antes do exame

```{r}

protf |> 
  dplyr::count(
    `22 - Paciente passou por cirurgia nos 30 dias que antecederam a realização do exame PROTF?`
  ) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

## h) Porcentagem de pacientes que passaram por tratamento oncológico nos 30 antes do exame

```{r}

protf |> 
  dplyr::count(
    `23 - Paciente passou por tratamento oncológico nos 30 dias que antecederam a realização do exame PROTF?`
  ) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

## i) Porcentagem de pacientes que fizeram uso de medicação

```{r}

protf |> 
  dplyr::count(
    `26 - Paciente faz uso de medicação que pode causar imunodeficiência nos 30 dias que antecederam a realização do exame PROTF?`
  ) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

## j) Porcentagem de pacientes com registro de óbito. Definir intervalo de tempo entre a data do exame e a data do óbito.

```{r}

protf |> 
  dplyr::count(`33 - Paciente possui registro de óbito?`) |>
  dplyr::mutate(Percentual = n / nrow(protf))

proxy <- 
  protf |> 
  dplyr::filter(`33 - Paciente possui registro de óbito?` == "Sim") |>
  dplyr::select(`1 - Insira o Prontuário do Paciente`, # ---------------------
                `5 - Data de Cadastro do Exame`, 
                `33.2 - Data de óbito`) |>
  dplyr::mutate(
    Intervalo_dias = lubridate::time_length(
      lubridate::interval(`5 - Data de Cadastro do Exame`, `33.2 - Data de óbito`), 
      "days"
    ),
    Intervalo_meses = lubridate::time_length(
      lubridate::interval(`5 - Data de Cadastro do Exame`, `33.2 - Data de óbito`), 
      "months"
    )
  )

proxy |> dplyr::select(-`1 - Insira o Prontuário do Paciente`)

summary(proxy$Intervalo_dias)
summary(proxy$Intervalo_meses)

```

146 pacientes vieram a óbito (18.29%), com um tempo médio de sobrevida de 193 dias (6.3 meses) - mediana de 40 dias (1.3 meses). 

# 2. Análise de indivíduos com dosagem de IgG

## a) A partir dos valores de dosagem de IgG, verificar se existe correlação com os valores de PROTF; Não considerar na estatistica os pacientes que fazem reposição de IgG - pois é um tratamento, logo não é um dado basal

138 pacientes (17.3%) tem registro do valor de IgG e não fizeram reposição.

```{r}

(
  proxy <- 
    protf |>
    dplyr::filter(
      `19 - Fez reposições de Imunoglobulinas durante os 30 dias que antecederam a realização do exame PROTF  ?`
    != "Sim" &
    `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica` != "NA"
    ) |> 
    dplyr::rename(IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`) |>
    dplyr::rename(PROTF = `16 - Resultado do Primeiro exame`) |>
    dplyr::select(IgG, PROTF) |>
    tidyr::unnest(IgG)  
)

```

Correlação linear de Pearson de 0.47 (forte e estatísticamente significativa, 6.6e-09 = 0.0000000066).

Contudo, vemos que o que melhor se adequa aos dados é um comportamento quadrático - subida linear (não tão forte) até um igG ao redor de 600 e a partir dali uma curva descendente, mas temos poucos pacientes passando desse valor.

```{r fig.width=8}

proxy |>
  ggplot() +
  aes(x = IgG, y = PROTF) +
  geom_point(size = 3, shape = 21, fill = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 2), 
              color = "red", 
              se = TRUE) + 
  ggpubr::stat_cor(method = "pearson", label.x = 1100, label.y = 1.3) + 
  labs(title = "Correlação entre IgG e PROTF",
       subtitle = "Modelo Linear Quadrático (y ~ x²) com banda de confiança de 95%",
       x = "IgG (mg/dL)", y = "PROTF (g/dL)") +
  scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
  scale_y_continuous(breaks = seq(0, 3, by = 0.1))

```

## b) Entre os indivíduos com dosagem de IgG abaixo do valor de referência, qual foi o valor médio, mínimo e máximo do valor de PROTF?

```{r fig.width=8}

protf |>
  dplyr::count(
    `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica`
    ) 

proxy <- 
  protf |>
    dplyr::filter(
    `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica` == "Abaixo do valor de referência"
    ) |> 
    dplyr::rename(IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`) |>
    dplyr::rename(PROTF = `16 - Resultado do Primeiro exame`) |>
    dplyr::select(IgG, PROTF) |>
    tidyr::unnest(IgG) 

summary(proxy$PROTF)

proxy |>
  ggplot() +
  aes(x = IgG, y = PROTF) +
  geom_point(size = 3, shape = 21, fill = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 1), 
              color = "red", 
              se = TRUE) + 
  ggpubr::stat_cor(method = "pearson", label.x = 400, label.y = 1.3) + 
  labs(title = "Correlação entre IgG e PROTF",
       subtitle = "Modelo Linear (y ~ x) com banda de confiança de 95%",
       x = "IgG (mg/dL)", y = "PROTF (g/dL)") +
  scale_x_continuous(breaks = seq(0, 550, by = 50)) +
  scale_y_continuous(breaks = seq(0, 3, by = 0.1))
  
```

PROTF médio de 1.744, min 1.2 e máx 2.4.

## c) Entre os indivíduos com dosagem de IgG acima do valor de referência e que NÃO fazem reposição de IgG, qual foi o valor médio, mínimo e máximo do valor de PROTF?

Nenhum paciente com IgG acima do valor de referência.

## d) Entre os indivíduos com dosagem de IgG abaixo do valor de referência, quantos possuíam comorbidade? Quais eram mais frequentes?

71 pacientes, 35 diferentes combinações de comorbidades - apenas 5 ocorrendo em ao menos dois pacientes.

```{r}

protf |>
  dplyr::filter(
    `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica` == "Abaixo do valor de referência"
  ) |>
  dplyr::count(`20 - Paciente possui alguma comorbidade/diagnostico de base?`)

protf |>
  dplyr::filter(
    `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica` == "Abaixo do valor de referência" &
  `20 - Paciente possui alguma comorbidade/diagnostico de base?` == "Sim"
  ) |>
  dplyr::count(`20.1 -  Qual comorbidade o paciente possui?

-Em caso negativo responder com NA`) |>
  dplyr::arrange(dplyr::desc(n)) |>
  dplyr::filter(n > 1)

```

Reunindo os resultados similares em um único grupo, temos 25 diferentes.

```{r}

protf |> 
  dplyr::filter(
        `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica` == "Abaixo do valor de referência" &
    `20 - Paciente possui alguma comorbidade/diagnostico de base?` == "Sim"
  ) |>
  dplyr::select(Comorbidades = `20.1 -  Qual comorbidade o paciente possui?

-Em caso negativo responder com NA`) |>
  tidyr::separate_longer_delim(cols = Comorbidades, delim = ";") |>
  dplyr::mutate(
    Comorbidades = stringr::str_trim(Comorbidades),
    Comorbidades = stringi::stri_trans_general(Comorbidades, "Latin-ASCII"),
    Comorbidades = stringr::str_to_sentence(Comorbidades)
  ) |>
  dplyr::count(Comorbidades) |> 
  dplyr::arrange(dplyr::desc(n)) |> 
  print(n = 30)
  
```

# 3. Análise dos menores de 1 ano

> Entre os menores de 1 ano, quantos apresentavam a comorbidade "Prematuridade"?

Dentre os 198 pacientes menores de 1 ano e com comorbidade, temos 78 diferentes combinações de comorbidades.

```{r}

protf |>
  dplyr::count(`3 - Insira Faixa etária do paciente`)

protf |>
  dplyr::filter(`3 - Insira Faixa etária do paciente` == "< 1 ano") |>
  dplyr::count(`20 - Paciente possui alguma comorbidade/diagnostico de base?`)

# protf |>
#   dplyr::filter(
#     `3 - Insira Faixa etária do paciente` == "< 1 ano"
#   ) |>
#     dplyr::rename(Comorbidade = `20.1 -  Qual comorbidade o paciente possui?
# 
# -Em caso negativo responder com NA`) |>
#   dplyr::count(Comorbidade)

protf |>
  dplyr::rename(Comorbidade = `20.1 -  Qual comorbidade o paciente possui?

-Em caso negativo responder com NA`) |>
  dplyr::filter(
    `3 - Insira Faixa etária do paciente` == "< 1 ano" &
    stringr::str_detect(Comorbidade, "Prematuridade")
  ) |>
  dplyr::count(Comorbidade) |>
  dplyr::arrange(dplyr::desc(n))

```

Destas 78, 20 apresentam "Prematuridade".

Somados:

```{r}

protf |> 
  dplyr::filter(
    `3 - Insira Faixa etária do paciente` == "< 1 ano" &
    `20 - Paciente possui alguma comorbidade/diagnostico de base?` == "Sim"
  ) |>
  dplyr::select(Comorbidades = `20.1 -  Qual comorbidade o paciente possui?

-Em caso negativo responder com NA`) |>
  tidyr::separate_longer_delim(cols = Comorbidades, delim = ";") |>
  dplyr::mutate(
    Comorbidades = stringr::str_trim(Comorbidades),
    Comorbidades = stringi::stri_trans_general(Comorbidades, "Latin-ASCII"),
    Comorbidades = stringr::str_to_sentence(Comorbidades)
  ) |>
  dplyr::count(Comorbidades) |> 
  dplyr::arrange(dplyr::desc(n)) |> 
  dplyr::filter(Comorbidades == "Prematuridade")
  
```

# Levando em consideração pacientes que fizeram os exames PROTF + IgG, definir um valor de referência para PROTF levando como referência a dosagem de IgG

147 pacientes.

```{r fig.width=8.5}

(
  proxy <- 
    protf |>
    dplyr::rename(IgG_Referência = `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica`) |>
    dplyr::rename(IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`) |>
    dplyr::rename(PROTF = `16 - Resultado do Primeiro exame`) |>
    dplyr::filter(IgG_Referência != "NA") |> 
    dplyr::select(IgG, PROTF, IgG_Referência) |>
    tidyr::unnest(IgG)  
)

proxy |> dplyr::count(IgG_Referência)

proxy |>
  ggplot() +
  aes(x = IgG, y = PROTF, fill =IgG_Referência, color = IgG_Referência) +
  geom_point(size = 3, shape = 21, alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 1), 
              color = "red", 
              se = TRUE) + 
  ggpubr::stat_cor(method = "pearson", 
                   label.x.npc = 0.7,   # 70% across the x-axis
                   label.y.npc = "bottom") + 
  labs(title = "Correlação entre IgG e PROTF",
       subtitle = "Modelo Linear (y ~ x) com banda de confiança de 95%",
       fill = "Referência: ", color = "Referência: ",
       x = "IgG (mg/dL)", y = "PROTF (g/dL)") +
  scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
  scale_y_continuous(breaks = seq(1, 3, by = 0.1))

(
  ref_stats <- proxy |>
    filter(IgG_Referência == "Normal") |>
    summarise(
      mean_protf = mean(PROTF),
      sd_protf = sd(PROTF),
      upper_ref = mean_protf + (2 * sd_protf),
      lower_ref = mean_protf - (2 * sd_protf)
    )
)

```

Valor de referência: 2.02 (95% CI 1.49, 2.55).

```{r fig.width=9, fig.height=6}

lin_mod <- lm(PROTF ~ IgG, data = proxy)

# Fit a segmented model (it estimates where the 'plateau' starts)
# You provide an initial guess for the breakpoint (e.g., IgG = 600)
seg_mod <- segmented(lin_mod, seg.Z = ~IgG, psi = 600)

# seg_mod$psi[, "Est."] + c(-1.95, 1.96) * seg_mod$psi[, "St.Err"]

proxy |>
  ggplot(aes(x = IgG, y = PROTF)) +
  annotate("rect", xmin = -Inf, xmax = Inf, 
           ymin = ref_stats$lower_ref, ymax = ref_stats$upper_ref, 
           fill = "steelblue", alpha = 0.3) +
  annotate("segment", x = -Inf, xend = Inf, 
           y = ref_stats$mean_protf, yend = ref_stats$mean_protf, 
           color = "steelblue", linetype = "dashed", linewidth = 1) +
  geom_point(aes(fill = IgG_Referência), shape = 21, size = 3, alpha = 0.8) +
  geom_smooth(
    aes(color = IgG_Referência), 
    method = "lm", formula = y ~ poly(x, 1), se = TRUE, linewidth = 1
  ) +
  geom_vline(xintercept = seg_mod$psi[, "Est."], linetype = "dashed", alpha = 0.7) +
  scale_fill_manual(
    values = c("Normal" = "white", "Abaixo do valor de referência" = "orange"),
    labels = c("Abaixo da Referência", "Normal")
  ) +
  scale_color_manual(
    values = c("Normal" = "darkblue", "Abaixo do valor de referência" = "darkorange"),
    labels = c("Abaixo da Referência", "Normal")
  ) +
  labs(
    title = "Baseline de PROTF Baseado no Platô de IgG (Grupo Normal)",
    subtitle = "Linhas de tendência com Intervalo de Confiança de 95% (sombreado)",
    x = "IgG (mg/dL)", 
    y = "PROTF (g/dL)",
    fill = "Classificação IgG: ", color = "Classificação IgG: "
  ) +
  scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
  scale_y_continuous(breaks = seq(1, 2.5, by = 0.1))

```

Além disso, estimamos um ponto de corte para ver onde se inicia o platô - onde termina o padrão de crescimento e se inicia a estagnação de PROTF médio em relação ao IgG, é em 616.

```{r}

summary(seg_mod)

```

Aqui é apenas para "comprovar" que o grupo "Normal" não tem inclinação estatísticamente significativa. Neste grupo, independente do valor de IgG se espera o mesmo valor de PROTF.

```{r}

proxy |>
  dplyr::filter(IgG_Referência != "Acima do valor de referência") |>
  dplyr::group_by(IgG_Referência) |>
  dplyr::group_modify(~ broom::tidy(lm(PROTF ~ IgG, data = .x), conf.int = TRUE)) |>
  dplyr::filter(term == "IgG") |>
  dplyr::select(IgG_Referência, estimate, std.error, p.value, conf.low, conf.high)

```

## Explicação mais detalhada

### 1. Establishment of the Baseline Reference

To define a physiological PROTF baseline, the Normal Group (patients with IgG within the reference range) was used as the normative standard.

* **Process:** We calculated the mean and standard deviation (SD) for the Normal group. The shaded area in the visualization represents the Mean ± 2 SD, covering approximately 95% of the expected physiological variation.

* **Interpretation:** PROTF values falling below this band indicate a pathological deficit relative to the established biological baseline.

### 2. Group Dynamics: The "Rise vs. Plateau" Effect

The relationship was modeled separately for each group to identify how IgG deficiency impacts PROTF measurements.

* **Below Reference Group:** This group exhibits a significant positive slope (p < 0.05). This demonstrates that in patients with low IgG, PROTF is "coupled" to IgG levels—as IgG increases, PROTF rises proportionally.

* **Normal Group:** This group shows a slope near zero (the confidence interval includes zero). This confirms a Plateau effect, where PROTF reaches a steady state and no longer responds to further IgG increases, representing the true biological baseline.

### 3. Segmented Regression (Inflection Point)

A Segmented (Broken-stick) Regression was applied to mathematically determine the threshold of stabilization.

* **The Breakpoint:** The model identifies the specific IgG coordinate where the linear rise ends and the plateau begins. 

* **Clinical Utility:** This point defines the "Critical IgG Threshold" required for a PROTF measurement to be considered stable and diagnostically reliable, independent of hypogammaglobulinemia interference.

---

### Summary of Statistical Findings

A positive linear dependency was observed between PROTF and IgG strictly within the deficient group (p < 0.05). Conversely, the normal group exhibited a plateau (p > 0.05), establishing a stable reference baseline.

# How to use PROTF to predict IgG deficiency

> Minimizing false positives, prioritizing Specificity and Positive Predictive Value (PPV).
>
> - Catch fewer sick people (lower sensitivity), but almost everyone catched will actually be sick.

```{r}

# 1. Prepare Data
proxy_roc <- proxy |>
  filter(IgG_Referência != "Acima do valor de referência") |>
  mutate(low_igg_flag = ifelse(IgG_Referência == "Abaixo do valor de referência", 1, 0))

# 2. Generate ROC Object
roc_obj <- roc(proxy_roc$low_igg_flag, proxy_roc$PROTF, quiet = TRUE)

# 3. Extract Scenarios

# A. Max Specificity (The "Perfect" but useless one)
scen_max_spec <- coords(roc_obj, "all", ret = c("threshold", "specificity", "sensitivity", "ppv")) |>
  filter(specificity == 1) |>
  slice(1) |>
  mutate(Strategy = "Max Specificity (100%)")

# B. High Specificity (The >90% attempt)
scen_high_spec <- coords(roc_obj, "all", ret = c("threshold", "specificity", "sensitivity", "ppv")) |>
  filter(specificity >= 0.90) |>
  arrange(desc(sensitivity)) |> # Get best sensitivity in this range
  slice(1) |>
  mutate(Strategy = "High Specificity (>90%)")

# C. Balanced (Youden's Index - The Winner)
scen_balanced <- coords(roc_obj, "best", best.method = "youden", 
                        ret = c("threshold", "specificity", "sensitivity", "ppv")) |>
  mutate(Strategy = "Balanced (Youden's Index)")

# 4. Combine into a Summary Table
roc_summary <- bind_rows(scen_max_spec, scen_high_spec, scen_balanced) |>
  dplyr::select(
    Strategy, 
    Threshold = threshold, 
    Sensitivity = sensitivity, Specificity = specificity, PPV = ppv
  )

# Display Table
kable(roc_summary, digits = 3, caption = "Comparison of Cutoff Strategies")

# 5. Visualization: Density Plot with the Winner Cutoff
optimal_cutoff <- scen_balanced$threshold

proxy |>
  filter(IgG_Referência != "Acima do valor de referência") |>
  ggplot(aes(x = PROTF, fill = IgG_Referência)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = optimal_cutoff, linetype = "dashed", color = "black", linewidth = 1) +
  annotate("text", x = optimal_cutoff, y = 0, label = paste(" Optimal Cutoff:", round(optimal_cutoff, 2)), 
           hjust = 1.05, vjust = -2, angle = 0, fontface = "bold") +
  scale_fill_manual(values = c("Normal" = "steelblue", "Abaixo do valor de referência" = "orange")) +
  labs(
    title = "Clinical Decision Threshold",
    subtitle = "Overlap of PROTF distributions with optimal screening cutoff",
    x = "PROTF (g/dL)", y = "Density", fill = "IgG Status"
  ) +
  scale_x_continuous(breaks = seq(1, 2.5, by = 0.1))

```

## Diagnostic Screening: PROTF as a Predictor for IgG Testing

While previous sections established how IgG influences PROTF, we also investigated the reverse relationship: identifying a PROTF threshold that serves as a clinical trigger for ordering an IgG quantification.

To determine this threshold, we performed a Receiver Operating Characteristic (ROC) curve analysis. We evaluated three distinct cutoff strategies to balance the trade-off between **Precision** (minimizing unnecessary tests) and **Safety** (preventing missed diagnoses).

### Analysis of Cutoff Strategies

The table above summarizes the performance of three potential thresholds:

1.  **Max Specificity Strategy (Cutoff ~1.45):**
    * **Pros:** Zero false positives (PPV = 100%).
    * **Cons:** Extremely low sensitivity (~11%). This threshold misses 9 out of 10 deficient patients, rendering it useless for screening.

2.  **High Specificity Strategy (Cutoff ~1.65):**
    * **Pros:** High precision (PPV > 80%).
    * **Cons:** Sensitivity remains poor (~32%). Using this cutoff would still fail to identify the majority of patients with hypogammaglobulinemia.

3.  **Balanced "Youden" Strategy (Cutoff ~2.05):**
    * **Pros:** Optimizes the sum of sensitivity and specificity. It achieves a **Sensitivity of ~87%**, ensuring that the vast majority of deficient patients are flagged.
    * **Cons:** Lower specificity (~54%), implying that roughly 1 in 3 positive screens will result in a normal IgG upon confirmation (False Positives).

### Conclusion and Recommendation

Based on the analysis, strictly minimizing false positives results in an unacceptable loss of diagnostic sensitivity. Therefore, we recommend the **Balanced Strategy** as the clinical standard.

**Recommendation:** A PROTF value below **2.05 g/L** should trigger a follow-up IgG test. This threshold prioritizes patient safety by maintaining high detection rates (Sensitivity ~87%), accepting a moderate rate of confirmatory testing as a necessary trade-off for an effective screening tool.

### Advanced Strategy: The "Indeterminate Zone" Approach

```{r}

# 1. Prepare Data & ROC
proxy_roc <- proxy |>
  filter(IgG_Referência != "Acima do valor de referência") |>
  mutate(low_igg_flag = ifelse(IgG_Referência == "Abaixo do valor de referência", 1, 0))

roc_obj <- roc(proxy_roc$low_igg_flag, proxy_roc$PROTF, quiet = TRUE)

# 2. Extract All Coordinates
all_coords <- coords(roc_obj, "all", ret = c("threshold", "specificity", "sensitivity"))

# 3. Find the Boundaries (Using >= Logic to avoid NA)

# A. LOWER CUTOFF (Red Zone Boundary)
# Logic: We want High Specificity (>= 95%) to be sure they are sick.
# We pick the threshold with the highest sensitivity possible within that constraint.
cutoff_red <- all_coords |>
  filter(specificity >= 0.95) |>
  arrange(desc(sensitivity)) |> 
  slice(1) |>
  pull(threshold)

# B. UPPER CUTOFF (Green Zone Boundary)
# Logic: We want High Sensitivity (>= 95%) to be sure we don't miss anyone.
# We pick the threshold with the highest specificity possible within that constraint.
cutoff_green <- all_coords |>
  filter(sensitivity >= 0.95) |>
  arrange(desc(specificity)) |> 
  slice(1) |>
  pull(threshold)

# Check if boundaries were found (Assign default if data is too sparse)
if(length(cutoff_red) == 0) cutoff_red <- min(proxy_roc$PROTF)
if(length(cutoff_green) == 0) cutoff_green <- max(proxy_roc$PROTF)

# 4. Print Values for verification
rbind(
  paste("Red Zone (Deficient) <", round(cutoff_red, 2)),
  paste("Grey Zone (Indeterminate):", round(cutoff_red, 2), "-", round(cutoff_green, 2)),
  paste("Green Zone (Safe) >", round(cutoff_green, 2))
)

```

```{r fig.width=9.5, fig.height=6}

# 2. Visualize the Zones
proxy |>
  filter(IgG_Referência != "Acima do valor de referência") |>
  ggplot(aes(x = PROTF, fill = IgG_Referência)) +
  geom_density(alpha = 0.4) +
  # Red Zone
  annotate("rect", xmin = -Inf, xmax = cutoff_red, ymin = 0, ymax = Inf, 
           fill = "firebrick", alpha = 0.1) +
  # Grey Zone
  annotate("rect", xmin = cutoff_red, xmax = cutoff_green, ymin = 0, ymax = Inf, 
           fill = "gold", alpha = 0.15) +
  # Green Zone
  annotate("rect", xmin = cutoff_green, xmax = Inf, ymin = 0, ymax = Inf, 
           fill = "forestgreen", alpha = 0.1) +
  geom_vline(xintercept = c(cutoff_red, cutoff_green), linetype = "dashed") +
  annotate("text", x = cutoff_red, y = 0, label = paste("Red <", round(cutoff_red, 2)), 
           hjust = 1.1, vjust = -2, angle = 0, color = "firebrick") +
  annotate("text", x = cutoff_green, y = 0, label = paste("Green >", round(cutoff_green, 2)), 
           hjust = 1.1, vjust = -2, angle = 0, color = "forestgreen") +
  scale_fill_manual(values = c("Normal" = "steelblue", "Abaixo do valor de referência" = "orange")) +
  labs(title = "Traffic Light System: Managing Clinical Overlap",
       subtitle = "Red: High Probability of Deficiency | Grey: Indeterminate | Green: High Probability of Normal",
       x = "PROTF (g/dL)", y = "Density", fill = "IgG Status") +
  scale_x_continuous(breaks = seq(1, 2.5, by = 0.1)) +
  scale_y_continuous(breaks = seq(0, 2, by = 0.25))

```

Given the significant distributional overlap between the IgG groups, a single binary cutoff forces a trade-off that compromises either diagnostic sensitivity or specificity. To address this biological reality, we applied a **Two-Cutoff Strategy** (Traffic Light System) to isolate the zone of diagnostic uncertainty.

We defined three clinical zones based on strict statistical safety margins:

1.  **Deficiency Zone (< 1.55 g/L):**
    * **Definition:** Threshold established at **≥95% Specificity**.
    * **Implication:** Results in this range are highly indicative of IgG deficiency. The probability of a false positive is minimal (<5%), warranting immediate clinical action.

2.  **Safety Zone (> 2.3 g/L):**
    * **Definition:** Threshold established at **≥95% Sensitivity**.
    * **Implication:** Results in this range effectively rule out deficiency with high confidence. The risk of missing a true deficient case is minimal (<5%).

3.  **Indeterminate "Grey" Zone (1.55 – 2.3 g/L):**
    * **Definition:** The interval between the high-specificity and high-sensitivity thresholds.
    * **Implication:** This range captures the patients whose PROTF values are inconclusive due to biological overlap. For these individuals, PROTF alone is insufficient, and **mandatory IgG quantification** is recommended to resolve the diagnosis.

**Conclusion:** By utilizing this tiered approach, we increase the diagnostic accuracy for patients at the extremes while clearly identifying the subset of patients for whom confirmatory testing is strictly necessary.

## Infections

```{r}

# Define your Cutoff (Use the Balanced one found earlier)
protf_cutoff <- 2.05 

infection_analysis <- protf |>
  dplyr::rename(PROTF = `16 - Resultado do Primeiro exame`) |>
  # Create Binary Groups based on Cutoff
  mutate(Risk_Group = ifelse(PROTF < protf_cutoff, "High_Risk", "Low_Risk")) |>
  dplyr::select(
    Risk_Group,
    `8 - Infecção em sistema nervoso central?`,
    `9 - Infecção de vias áreas superiores?`,
    `10 - Infecção de vias áreas inferiores?`,
    `11 - Infecção de sistema urinário?`,
    `12 - Infecção de sistema gastrointestinal?`,
    `13 - Infecção de pele?`,
    `14 - Sepse?`,
    `15 - Outros?`
  ) |>
  pivot_longer(
    cols = !Risk_Group,
    names_to = "Infection_Type",
    values_to = "Status"
  ) |>
  mutate(
    Infection_Type = stringr::str_remove(Infection_Type, "^\\d+ - "),
    Infection_Type = stringr::str_remove(Infection_Type, "\\?"),
    Infection_Type = stringr::str_remove(Infection_Type, "^Infecção (de|em) "),
    Infection_Type = stringr::str_to_sentence(Infection_Type),
    Is_Infected = ifelse(Status == "Sim", 1, 0)
  ) |>
  # Remove NAs if any exist in Status/Risk_Group
  filter(!is.na(Is_Infected))

rates_table <- infection_analysis |>
  group_by(Infection_Type, Risk_Group) |>
  summarise(
    Total_Patients = n(),
    Infected_Count = sum(Is_Infected),
    Rate_Pct = (Infected_Count / Total_Patients) * 100, # Em porcentagem
    .groups = "drop"
  ) |>
  # Coloca High_Risk e Low_Risk lado a lado
  pivot_wider(
    names_from = Risk_Group,
    values_from = c(Total_Patients, Infected_Count, Rate_Pct)
  )

stats_table <- infection_analysis |>
  group_by(Infection_Type) |>
  nest() |>
  mutate(
    model = map(data, function(df) {
      # Tabela cruzada 2x2
      tbl <- table(df$Risk_Group, df$Is_Infected)
      
      # Verifica se a tabela é válida (tem 2 grupos de risco e pelo menos infectados e não infectados no geral)
      # Se for muito esparsa, fisher.test pode falhar ou retornar NULL
      tryCatch({
        broom::tidy(fisher.test(tbl))
      }, error = function(e) NULL)
    })
  ) |>
  unnest(model) |>
  dplyr::select(Infection_Type, estimate, p.value, conf.low, conf.high)

final_results <- rates_table |>
  left_join(stats_table, by = "Infection_Type") |>
  mutate(
    Odds_Ratio = estimate,
    Significant = ifelse(p.value < 0.05, "Yes", "No")
  ) |>
  dplyr::select(
    Infection_Type, 
    Rate_High = Rate_Pct_High_Risk, 
    Rate_Low = Rate_Pct_Low_Risk,
    Odds_Ratio, 
    conf.low, 
    conf.high,
    p.value, 
    Significant
  )

knitr::kable(
  final_results, digits = 2, 
  caption = "Prevalência de Infecções por Grupo de Risco (PROTF < 2.05)"
)

```

Odds Ratio: Normal (>= 2.05) vs. Low (< 2.05). 0.3 means que o grupo normal tem menos chances.

```{r fig.width=8, fig.height=4}

final_results |>
  filter(!is.na(Odds_Ratio)) |> 
  ggplot(aes(x = Odds_Ratio, y = reorder(Infection_Type, Odds_Ratio), color = Significant)) +
  geom_point(size = 3.5) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.3, linewidth = 0.8) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray40", linewidth = 1) +
  scale_x_log10() + 
  scale_color_manual(values = c("Yes" = "#D55E00", "No" = "gray60")) +
  labs(
    title = "Clinical Impact of Low PROTF (< 2.05 g/dL)",
    subtitle = "Odds Ratio of Infection Risk (High Risk vs. Low Risk Group)",
    x = "Odds Ratio (Log Scale) \n >1 indicates higher risk in Low PROTF group",
    y = NULL,
    color = "Significant (p < 0.05)?"
  )

```

### Clinical Validation: Association with Infection History

To validate the clinical relevance of the established cutoff (**2.05 g/L**), we assessed whether patients falling below this threshold exhibited a distinct profile of infectious history compared to those above it.

* **Methodology:** Patients were stratified into "High Risk" (PROTF < 2.05) and "Low Risk" (PROTF ≥ 2.05) cohorts. We calculated the Prevalence Rate (%) and Odds Ratio (OR) for seven specific infection types using Fisher's Exact Test.

* **Results:**

    * **Severe Systemic Vulnerability (Sepsis):** The most critical finding was observed for Sepsis. The "High Risk" (Low PROTF) cohort exhibited a dramatically higher prevalence of sepsis (**42.62%**) compared to the "Low Risk" cohort (**18.42%**). The Odds Ratio of **0.3** (for the Low Risk group) confirms that maintaining PROTF levels above the cutoff is a significant protective factor against life-threatening systemic infections ($p < 0.001$).

    * **Distinct Phenotypes (Mucosal Infections):** Interestingly, the "Low Risk" (Normal PROTF) group reported significantly higher rates of common mucosal infections compared to the deficient group. This included **Upper Respiratory Tract Infections** (40.53% vs 18.58%, $p < 0.001$) and **Gastrointestinal Infections** (21.05% vs 4.37%, $p < 0.001$). This suggests distinct clinical phenotypes: patients with low PROTF are prone to severe complications (Sepsis), whereas those with normal levels present with more frequent, but likely less lethal, common community-acquired infections.

    * **Neutral Associations:** No statistically significant difference was found for Skin, CNS, Urinary, or Lower Respiratory infections ($p > 0.05$), indicating that the cutoff is highly specific to the risk of Sepsis rather than a generalized increase in all infection types.

**Conclusion:** The PROTF cutoff of 2.05 g/L successfully identifies a patient subset with critical immunological vulnerability. While patients above this threshold frequently report common mucosal conditions, those falling below it are disproportionately susceptible to severe sepsis, validating the threshold as a vital marker for systemic risk stratification.

## By Age Interval

```{r fig.width=9.75, fig.height=7}

# protf |>
#   dplyr::filter(`3 - Insira Faixa etária do paciente` != "> 18 anos") |>
#   dplyr::count(`3 - Insira Faixa etária do paciente`)

proxy <- 
    protf |>
    dplyr::rename(IgG_Referência = `18.1 - Como se encontra o resultado de IgG?

NA = não se aplica`) |>
    dplyr::rename(IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`) |>
    dplyr::rename(PROTF = `16 - Resultado do Primeiro exame`) |>
    dplyr::rename(Age_Group = `3 - Insira Faixa etária do paciente`) |>
    dplyr::filter(IgG_Referência != "NA") |> 
    dplyr::select(IgG, PROTF, IgG_Referência, Age_Group) |>
    tidyr::unnest(IgG)  

# 1. Prepare Data
age_analysis_data <- proxy |>
  dplyr::filter(
    IgG_Referência != "Acima do valor de referência",
    Age_Group != "> 18 anos" 
  ) |>
  dplyr::mutate(low_igg_flag = ifelse(IgG_Referência == "Abaixo do valor de referência", 1, 0))

# 2. Define a function to calculate scenarios for a single group
calc_roc_scenarios <- function(data) {
  # Safety check: Need both 0s and 1s to make an ROC curve
  if(n_distinct(data$low_igg_flag) < 2) return(NULL)
  
  # Generate ROC
  roc_obj <- roc(data$low_igg_flag, data$PROTF, quiet = TRUE)
  
  # A. Max Specificity
  s1 <- coords(roc_obj, "all", ret = c("threshold", "specificity", "sensitivity", "ppv")) |>
    dplyr::filter(specificity == 1) |> 
    dplyr::slice(1) |> 
    dplyr::mutate(Strategy = "Max Spec (100%)")
  
  # B. High Specificity
  s2 <- coords(roc_obj, "all", ret = c("threshold", "specificity", "sensitivity", "ppv")) |>
    dplyr::filter(specificity >= 0.90) |> 
    dplyr::arrange(desc(sensitivity)) |> 
    dplyr::slice(1) |> 
    dplyr::mutate(Strategy = "High Spec (>90%)")
  
  # C. Balanced (Youden)
  s3 <- coords(roc_obj, "best", best.method = "youden", 
               ret = c("threshold", "specificity", "sensitivity", "ppv")) |> 
    dplyr::mutate(Strategy = "Balanced (Youden)")
  
  dplyr::bind_rows(s1, s2, s3)
}

# 3. Apply function to each Age Group
stratified_results <- age_analysis_data |>
  dplyr::group_by(Age_Group) |>
  tidyr::nest() |>
  dplyr::mutate(roc_stats = map(data, calc_roc_scenarios)) |>
  tidyr::unnest(roc_stats) |>
  dplyr::select(
    Age_Group, Strategy, 
    Threshold = threshold, 
    Sensitivity = sensitivity, Specificity = specificity, PPV = ppv
  ) |>
  dplyr::arrange(Age_Group, Strategy)

# Display Table
kable(
  stratified_results |>
  dplyr::inner_join(proxy |> dplyr::count(Age_Group), by = "Age_Group") |>
  dplyr::relocate(n, .after = Age_Group), 
  digits = 3, caption = "Optimal PROTF Cutoffs by Age Group"
)

# 4. Visualization: Faceted Density Plot
# First, extract the "Balanced" cutoff for each group to draw the lines
balanced_cutoffs <- stratified_results |>
  dplyr::filter(Strategy == "Balanced (Youden)") |>
  dplyr::select(Age_Group, cutoff = Threshold)

age_analysis_data |>
  ggplot(aes(x = PROTF, fill = IgG_Referência)) +
  geom_density(alpha = 0.4) +
  # Add the specific cutoff line for each facet
  geom_vline(data = balanced_cutoffs, aes(xintercept = cutoff), 
             linetype = "dashed", color = "black", linewidth = 0.8) +
  # Add text label for the cutoff
  geom_text(data = balanced_cutoffs, aes(x = cutoff, y = 0, label = round(cutoff, 3)), 
            inherit.aes = FALSE, vjust = -1, hjust = -0.25, angle = 0, fontface = "bold") +
  facet_wrap(~Age_Group, scales = "free_y") + # Create separate panels
  scale_fill_manual(values = c("Normal" = "steelblue", "Abaixo do valor de referência" = "orange")) +
  labs(
    title = "PROTF Diagnostic Thresholds by Age Group",
    subtitle = "Distribution and Optimal Cutoff (Youden) per cohort",
    x = "PROTF (g/dL)", y = "Density", fill = "IgG Status"
  ) +
  scale_x_continuous(breaks = seq(1, 3, by = 0.1))

```

## Age-Stratified Analysis: Defining Precision Thresholds for PROTF

To determine if the global PROTF cutoff (2.05 g/L) is valid across all developmental stages, we performed a stratified ROC analysis. The results reveal critical variations in diagnostic utility, particularly regarding the test's inability to screen infants effectively.

### Results by Age Cohort

1.  **Infants (< 1 year): A Diagnostic Limitation**
    * **Optimal Cutoff:** 1.85 g/L (Statistical artifact).
    * **Performance:** The test demonstrates a **Sensitivity of only 3.7%**.
    * **Interpretation:** The analysis indicates that PROTF levels in IgG-deficient infants are virtually indistinguishable from healthy infants in this dataset. Consequently, **PROTF cannot be used to screen for IgG deficiency in the first year of life**, as it fails to detect >96% of cases.

2.  **Toddlers (1 to 3 years) & Adolescents (10 to 17 years):**
    * **Optimal Cutoff:** **2.05 g/L**.
    * **Performance:**
        * *Toddlers:* Sensitivity 61.5%, Specificity 66.7%.
        * *Adolescents:* Sensitivity 57.1%, Specificity 85.7%.
    * **Interpretation:** In these groups, the optimal threshold aligns perfectly with the global cutoff (2.05). However, diagnostic sensitivity is moderate (~60%), suggesting that while specific, the test may miss approximately 40% of deficient patients in these age ranges.

3.  **Children (4 to 9 years):**
    * **Optimal Cutoff:** **2.15 g/L**.
    * **Performance:** This group achieved the highest sensitivity (**75.0%**) among all cohorts.
    * **Interpretation:** The optimal threshold shifts slightly higher to 2.15 g/L. While detection rates improve in this demographic, the Positive Predictive Value (PPV) remains low (25.0%), indicating a higher rate of false positives compared to adolescents.

### Clinical Recommendation

Based on the age-stratified data, we propose the following protocols:

* **Contraindication (< 1 Year):** PROTF should **not** be used as a screening tool for infants under 1 year of age due to insufficient sensitivity.
* **Standard Cutoff (> 1 Year):** For all other age groups, a baseline cutoff of **2.05 g/L** is statistically supported.
* **Optimization (4-9 Years):** For children aged 4 to 9, clinicians may consider raising the threshold to **2.15 g/L** to maximize sensitivity (75%), though this comes with a trade-off of increased confirmatory testing (lower PPV).

```{r warning=FALSE, fig.width=9, fig.height=7}

# 1. Define Age-Specific Cutoffs for PROTF
# Based on your ROC results table
cutoff_map <- tibble(
  Age_Group = c("< 1 ano", "1 ano a 3 anos", "4 anos a 9 anos", "10 anos a 17 anos"),
  Age_Specific_Cutoff = c(1.85, 2.05, 2.15, 2.05)
)

# 2. Prepare Data & Apply Cutoffs
infection_analysis_age <- protf |>
  dplyr::rename(
    PROTF = `16 - Resultado do Primeiro exame`,
    Age_Group = `3 - Insira Faixa etária do paciente`
  ) |>
  # Filter to keep only the groups in our map
  dplyr::filter(Age_Group %in% cutoff_map$Age_Group) |>
  dplyr::left_join(cutoff_map, by = "Age_Group") |>
  
  # Create Risk Group based on the SPECIFIC cutoff for that age
  dplyr::mutate(Risk_Group = ifelse(PROTF < Age_Specific_Cutoff, "High_Risk", "Low_Risk")) |>
  
  # Select Infection Columns (8 to 14)
  dplyr::select(
    Age_Group,
    Risk_Group,
    matches("^(8|9|10|11|12|13|14|15) -")
  ) |>
  
  # Pivot Longer
  tidyr::pivot_longer(
    cols = matches("^(8|9|10|11|12|13|14|15) -"), 
    names_to = "Infection_Type",
    values_to = "Status"
  ) |>
  
  # Clean Names & Force Binary
  dplyr::mutate(
    Infection_Type = stringr::str_remove(Infection_Type, "^\\d+ - "),
    Infection_Type = stringr::str_remove(Infection_Type, "\\?"),
    Infection_Type = stringr::str_remove(Infection_Type, "^Infecção (de|em) "),
    Infection_Type = stringr::str_to_sentence(Infection_Type),
    Is_Infected = ifelse(Status == "Sim", 1, 0)
  ) |>
  dplyr::filter(!is.na(Is_Infected))

# 3. Calculate Rates (Stratified by Age AND Infection)
rates_table_age <- infection_analysis_age |>
  dplyr::group_by(Age_Group, Infection_Type, Risk_Group) |>
  dplyr::summarise(
    Total_Patients = n(),
    Infected_Count = sum(Is_Infected),
    Rate_Pct = (Infected_Count / Total_Patients) * 100,
    .groups = "drop"
  ) |>
  tidyr::pivot_wider(
    names_from = Risk_Group,
    values_from = c(Total_Patients, Infected_Count, Rate_Pct)
  )

# 4. Calculate Stats (Fisher Test within each Age/Infection combo)
stats_table_age <- infection_analysis_age |>
  dplyr::group_by(Age_Group, Infection_Type) |>
  tidyr::nest() |>
  dplyr::mutate(
    model = map(data, function(df) {
      tbl <- table(df$Risk_Group, df$Is_Infected)
      
      # Safety Check: Fisher test needs 2x2 dimensions
      tryCatch({
        broom::tidy(fisher.test(tbl))
      }, error = function(e) NULL)
    })
  ) |>
  tidyr::unnest(model) |>
  dplyr::select(Age_Group, Infection_Type, estimate, p.value, conf.low, conf.high)

# 5. Join and Finalize
final_results_age <- rates_table_age |>
  dplyr::left_join(stats_table_age, by = c("Age_Group", "Infection_Type")) |>
  dplyr::mutate(
    Odds_Ratio = estimate,
    Significant = ifelse(p.value < 0.05, "Yes", "No")
  ) |>
  dplyr::select(
    Age_Group,
    Infection_Type,
    Rate_High = Rate_Pct_High_Risk,
    Rate_Low = Rate_Pct_Low_Risk,
    Odds_Ratio,
    conf.low,
    conf.high,
    p.value,
    Significant
  ) |>
  dplyr::arrange(Age_Group, Infection_Type)

# Display Table
kable(final_results_age, digits = 2, caption = "Impact of Low PROTF on Infections (Stratified by Age)")

final_results_age |>
  filter(!is.na(Odds_Ratio)) |> # Hide rows where test failed (e.g. 0 infections)
  ggplot(aes(x = Odds_Ratio, y = reorder(Infection_Type, Odds_Ratio), color = Significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(xmin = conf.low, xmax = conf.high), width = 0.3) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  
  # Creates a separate panel for each age group
  facet_wrap(~Age_Group, scales = "free_x") + 
  
  scale_x_log10() +
  scale_color_manual(values = c("Yes" = "#D55E00", "No" = "gray60")) +
  labs(
    title = "PROTF Infection Risk by Age Cohort",
    subtitle = "Odds Ratio (High Risk vs Low Risk) using age-specific cutoffs",
    x = "Odds Ratio (Log Scale)",
    y = NULL
  )

```

### Age-Specific Clinical Validation: Infection Risk

To determine if the PROTF cutoffs serve as functional predictors of infection risk within specific age groups, we compared infection prevalence between "High Risk" (Below Cutoff) and "Low Risk" (Above Cutoff) patients.

* **Methodology:** Patients were stratified by age, applying the specific cutoffs: **1.85** (<1y), **2.05** (1-3y & 10-17y), and **2.15** (4-9y).

* **Results by Cohort:**

    1.  **Infants (< 1 year): Statistical Confirmation of Inutility**
        * **Observation:** The analysis yielded extremely wide confidence intervals (e.g., 0.00 to 1035.97) and p-values of 1.00 for most categories.
        * **The Cause:** The "Low Risk" (Control) group had **0.00%** infection rates across almost all categories. This is likely because the cutoff of 1.85 is statistically artificial for this group, failing to create a valid "healthy" control group for comparison.
        * **Conclusion:** This mathematically confirms the ROC conclusion: PROTF fails to discriminate between sick and healthy infants, rendering it useless as a clinical predictor for this age group.

    2.  **Children (4 to 9 years): Trends in Respiratory Health**
        * **Trend:** A visible clinical trend was observed for **Upper Respiratory Infections**. The "High Risk" group had a prevalence of **55.88%** compared to **39.47%** in the "Low Risk" group.
        * **Significance:** Despite the ~16% difference in raw rates, this did not reach statistical significance ($p = 0.15$), suggesting that while PROTF might correlate with respiratory susceptibility, a larger sample size is needed to confirm it.

    3.  **Toddlers (1-3y) and Adolescents (10-17y): Absence of Signal**
        * **Results:** No statistically significant associations were found. P-values for all infection types were > 0.05.
        * **Contrast with Global Analysis:** While the global analysis showed a strong link to Sepsis, this signal is lost in the subgroups. This suggests that the association with Sepsis is a "cumulative" signal that requires the statistical power of the full cohort to detect.

**Conclusion:** The age-stratified infection analysis reinforces the limitation of PROTF. It validates the exclusion of infants from screening. For older children, while trends exist (particularly for respiratory infections in the 4-9y group), PROTF should be viewed primarily as a biochemical marker for IgG deficiency rather than a standalone predictor of specific infection phenotypes.

# Medicações vs PROTF e IgG

```{r}

protf |> 
  dplyr::count(`34 -  Medicações Utilizadas`) |>
  dplyr::mutate(Percentual = n / nrow(protf)) |> dplyr::arrange(dplyr::desc(n))

protf |>
  tidyr::separate_longer_delim(
    cols = `34 -  Medicações Utilizadas`, delim = ";"
  ) |>
  dplyr::mutate(
    Medicações = stringi::stri_trans_general(
      `34 -  Medicações Utilizadas`, "Latin-ASCII"
    ),
    Medicações = stringr::str_trim(Medicações),    
    Medicações = stringr::str_to_sentence(Medicações)
  ) |>
  tidyr::drop_na(Medicações) |>
  dplyr::count(Medicações) |>
  dplyr::arrange(dplyr::desc(n))

```

Medidas descritivas de PROTF e IgG pelo uso de medicação. Aparentemente não vemos diferença significativa.

```{r}

protf |>
  dplyr::select(
    `34 -  Medicações Utilizadas`, 
    PROTF = `16 - Resultado do Primeiro exame`,
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA"))),
    Medicação = ifelse(
      is.na(`34 -  Medicações Utilizadas`), 
      "Com medicação", "Sem medicação"
    )
  ) |>
  dplyr::summarise(
    n_patients = dplyr::n(),
    dplyr::across(
      c(PROTF, IgG),
      list(
        Min = ~ min(.x, na.rm = TRUE),
        Q1 = ~ quantile(.x, 0.25, na.rm = TRUE),
        Median = ~ median(.x, na.rm = TRUE),
        Mean = ~ mean(.x, na.rm = TRUE),
        Q3 = ~ quantile(.x, 0.75, na.rm = TRUE),
        Max = ~ max(.x, na.rm = TRUE),
        NAs = ~ sum(is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    ),
    .by = Medicação
  ) |>
  tidyr::pivot_longer(
    cols = -Medicação, 
    names_to = "Metric", 
    values_to = "Value"
  ) |>
  tidyr::pivot_wider(names_from = Medicação, values_from = Value
  )

```

Graficamente fica claro que o fato de usar ou não medicamente não impacta nos valores de PROTF e nem de IgG, não existe influência / associação.

```{r fig.width=9, fig.height=6}

protf |>
  dplyr::select(
    `34 -  Medicações Utilizadas`, 
    PROTF = `16 - Resultado do Primeiro exame`,
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA"))),
    Medicação = ifelse(
      is.na(`34 -  Medicações Utilizadas`), 
      "Com medicação", "Sem medicação"
    )
  ) |>
  tidyr::drop_na(IgG) |>
  ggplot() +
  aes(x = IgG, y = PROTF, fill = Medicação, color = Medicação) +
  geom_point(size = 3, shape = 21, alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 1), 
              color = "red", 
              se = TRUE) + 
  ggpubr::stat_cor(method = "pearson", 
                   label.x.npc = 0.7,   # 70% across the x-axis
                   label.y.npc = "bottom") + 
  labs(title = "Correlação entre IgG e PROTF",
       subtitle = "Modelo Linear (y ~ x) com banda de confiança de 95%",
       fill = "Medicação: ", color = "Medicação: ",
       x = "IgG (mg/dL)", y = "PROTF (g/dL)") +
  scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
  scale_y_continuous(breaks = seq(1, 2.6, by = 0.1))

```

```{r fig.width=9, fig.height=4.5}

p1 <- protf |>
  dplyr::select(
    `34 -  Medicações Utilizadas`, 
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA"))),
    Medicação = ifelse(
      is.na(`34 -  Medicações Utilizadas`), 
      "Com medicação", "Sem medicação"
    )
  ) |>
  tidyr::drop_na(IgG) |>
  ggplot() +
  aes(x = IgG, fill = Medicação) +
  geom_histogram(alpha = 0.8, binwidth = 50) +
  labs(title = "Histograma de IgG",
       subtitle = "Binwidth: 50",
       fill = "Medicação: ",
       x = "IgG (mg/dL)", y = "n") +
  scale_x_continuous(breaks = seq(0, 1500, by = 200)) +
  scale_y_continuous(breaks = seq(0, 30, by = 5))
  
p2 <- protf |>
  dplyr::select(
    `34 -  Medicações Utilizadas`, 
    PROTF = `16 - Resultado do Primeiro exame`
  ) |>
  dplyr::mutate(
    Medicação = ifelse(
      is.na(`34 -  Medicações Utilizadas`), 
      "Com medicação", "Sem medicação"
    )
  ) |>
  ggplot() +
  aes(x = PROTF, fill = Medicação) +
  geom_histogram(alpha = 0.8, binwidth = 0.1) +
  labs(title = "Histograma de PROTF",
       subtitle = "Binwidth: 0.1",
       fill = "Medicação: ",
       x = "PROTF (g/dL)", y = "n") +
  scale_x_continuous(breaks = seq(1, 2.5, by = 0.2)) +
  scale_y_continuous(breaks = seq(0, 150, by = 15))

p1 + p2 + patchwork::plot_layout(guides = "collect")

```

Ajustamos também um modelo linear para entender melhor a associação entre IgG e o uso de medicação - após verificar os resíduos e constatar que o ajuste estava adequado, performamos uma Análise de variância (ANOVA) e com base no teste F obtivemos um valor-p de 0.3392, indicando que realmente não existe associação entre o valor de IgG e o fato de utilizar ou não um medicamento.

```{r}

lm(
  IgG ~ Medicação,
  data = protf |>
  dplyr::select(
    `34 -  Medicações Utilizadas`, 
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA"))),
    Medicação = ifelse(
      is.na(`34 -  Medicações Utilizadas`), 
      "Com medicação", "Sem medicação"
    )
  ) |>
  tidyr::drop_na(IgG)
) |>
  anova()

```

# Medicações mais prevalentes por Idade e Sexo

Abaixo mostramos - por Sexo.

```{r}

t1 <- protf |>
  dplyr::select(
    Medicações = `34 -  Medicações Utilizadas`,
    Idade = `3 - Insira Faixa etária do paciente`,
    Sexo = `4 - Sexo`
  ) |>
  tidyr::drop_na(Medicações) |>
  tidyr::separate_longer_delim(cols = Medicações, delim = ";") |>
  dplyr::mutate(Medicações = stringr::str_trim(Medicações)) |>
  dplyr::count(Medicações, Idade, Sexo) |>
  dplyr::arrange(dplyr::desc(n), Medicações) |>
  # dplyr::filter(n > 1) |>
  dplyr::group_split(Sexo) 

t1[[1]] |> print(n = 30)
t1[[2]] |> print(n = 30)

```

Mesma coisa pro faixa etária.

```{r}

protf |>
  dplyr::select(
    Medicações = `34 -  Medicações Utilizadas`,
    Idade = `3 - Insira Faixa etária do paciente`,
    Sexo = `4 - Sexo`
  ) |>
  tidyr::drop_na(Medicações) |>
  tidyr::separate_longer_delim(cols = Medicações, delim = ";") |>
  dplyr::mutate(Medicações = stringr::str_trim(Medicações)) |>
  dplyr::count(Medicações, Idade, Sexo) |>
  dplyr::arrange(dplyr::desc(n), Medicações, Sexo) |>
  # dplyr::filter(n > 1) |>
  dplyr::group_split(Idade)
  
```

E para finalizar, todas as medicações.

```{r}

protf |>
  dplyr::select(
    Medicações = `34 -  Medicações Utilizadas`
  ) |>
  tidyr::drop_na(Medicações) |>
  tidyr::separate_longer_delim(cols = Medicações, delim = ";") |>
  dplyr::mutate(Medicações = stringr::str_trim(Medicações)) |>
  dplyr::count(Medicações) |>
  dplyr::arrange(dplyr::desc(n)) |>
  print(n = 100)

```

# Perda proteica na urina (PPU)

> Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?

```{r}

protf |>
  dplyr::select(
    PPU=`24 - Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?`
  ) |>
  dplyr::count(PPU) |>
  dplyr::mutate(Percentual = n / nrow(protf))

```

Medidas descritivas.

```{r}

protf |>
  dplyr::select(
    PPU=`24 - Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?`,
    PROTF = `16 - Resultado do Primeiro exame`,
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA")))
  ) |>
  dplyr::summarise(
    n_patients = dplyr::n(),
    dplyr::across(
      c(PROTF, IgG),
      list(
        Min = ~ min(.x, na.rm = TRUE),
        Q1 = ~ quantile(.x, 0.25, na.rm = TRUE),
        Median = ~ median(.x, na.rm = TRUE),
        Mean = ~ mean(.x, na.rm = TRUE),
        Q3 = ~ quantile(.x, 0.75, na.rm = TRUE),
        Max = ~ max(.x, na.rm = TRUE),
        NAs = ~ sum(is.na(.x))
      ),
      .names = "{.col}_{.fn}"
    ),
    .by = PPU
  ) |>
  tidyr::pivot_longer(
    cols = -PPU, 
    names_to = "Metric", 
    values_to = "Value"
  ) |>
  tidyr::pivot_wider(names_from = PPU, values_from = Value
  )

```

Tanto pelas medidas quanto visualmente não aparenta ter diferença significativa tanto de PROTF quanto de IgG pelo fato de ter tido perda proteica na urina ou não.

```{r fig.width=9, fig.height=6}

protf |>
  dplyr::select(
    PPU=`24 - Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?`, 
    PROTF = `16 - Resultado do Primeiro exame`,
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA")))
  ) |>
  tidyr::drop_na(IgG) |>
  ggplot() +
  aes(x = IgG, y = PROTF, fill = PPU, color = PPU) +
  geom_point(size = 3, shape = 21, alpha = 0.6) +
  geom_smooth(method = "lm", 
              formula = y ~ poly(x, 1), 
              color = "red", 
              se = TRUE) + 
  ggpubr::stat_cor(method = "pearson", 
                   label.x.npc = 0.7,   # 70% across the x-axis
                   label.y.npc = "bottom") + 
  labs(title = "Correlação entre IgG e PROTF",
       subtitle = "Modelo Linear (y ~ x) com banda de confiança de 95%",
       fill = "Perda proteica na urina: ", color = "Perda proteica na urina: ",
       x = "IgG (mg/dL)", y = "PROTF (g/dL)") +
  scale_x_continuous(breaks = seq(0, 1500, by = 100)) +
  scale_y_continuous(breaks = seq(1, 2.6, by = 0.1))

```

```{r fig.width=9, fig.height=4.5}

p1 <- protf |>
  dplyr::select(
    PPU=`24 - Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?`, 
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA")))
  ) |>
  tidyr::drop_na(IgG) |>
  ggplot() +
  aes(x = IgG, fill = PPU) +
  geom_histogram(alpha = 0.8, binwidth = 50) +
  labs(title = "Histograma de IgG",
       subtitle = "Binwidth: 50",
       fill = "Perda proteica na urina: ",
       x = "IgG (mg/dL)", y = "n") +
  scale_x_continuous(breaks = seq(0, 1500, by = 200)) +
  scale_y_continuous(breaks = seq(0, 30, by = 5))
  
p2 <- protf |>
  dplyr::select(
    PPU = `24 - Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?`, 
    PROTF = `16 - Resultado do Primeiro exame`
  ) |>
  ggplot() +
  aes(x = PROTF, fill = PPU) +
  geom_histogram(alpha = 0.8, binwidth = 0.1) +
  labs(title = "Histograma de PROTF",
       subtitle = "Binwidth: 0.1",
       fill = "Perda proteica na urina: ",
       x = "PROTF (g/dL)", y = "n") +
  scale_x_continuous(breaks = seq(1, 2.5, by = 0.2)) +
  scale_y_continuous(breaks = seq(0, 150, by = 15))

p1 + p2 + patchwork::plot_layout(guides = "collect")

```

Ajustamos também um modelo linear para entender melhor a associação entre IgG e a perda proteica de urina - após verificar os resíduos e constatar que o ajuste estava adequado, performamos uma Análise de variância (ANOVA) e com base no teste F obtivemos um valor-p de 0.7344, indicando que realmente não existe associação entre o valor de IgG e o fato de ter tido perda proteica na urina.

```{r}

lm(
  IgG ~ PPU,
  data = protf |>
  dplyr::select(
    PPU = `24 - Paciente apresenta perda proteica na urina nos 30 dias que antecederam a realização do exame PROTF?`, 
    IgG = `18.2 - Qual resultado da dosagem? Colocar apenas número

-Responder com NA em caso de não realização do exame`
  ) |>
  dplyr::mutate(
    IgG = unlist(IgG),
    IgG = suppressWarnings(as.numeric(dplyr::na_if(IgG, "NA")))
  ) |>
  tidyr::drop_na(IgG)
) |>
  anova()

```

# Methodology & References

All analyses were conducted using the `R` language and environment for statistical computing (R Core Team, 2025). Data manipulation and string processing were performed using `dplyr` (Wickham et al., 2023), `tidyr` (Wickham et al., 2024), and `stringr` (Wickham, 2023). Temporal data were handled with `lubridate` (Grolemund & Wickham, 2011).

To date, the study has focused on data cleaning, feature engineering, and advanced visualization using `ggplot2` (Wickham, 2016). A key objective was the identification of biological cutoff points for PROTF where existing references were unknown or required validation. To achieve this, we utilized a two-step modeling approach:

1. **Reference Interval Modeling:** We established biological baselines by calculating the mean and standard deviation (± 2 SD) of the "Normal" IgG cohort, treating this group as a stable physiological standard.
2. **Threshold Discovery via Segmented Regression:** To identify the specific IgG level at which PROTF stabilize, we implemented segmented (broken-stick) regression (Muggeo, 2008). This allowed us to mathematically determine the "inflection point" between the rising deficient phase and the steady-state plateau phase.

Statistical significance of the observed slopes and the fit of the resulting plateaus were verified using tidied coefficients extracted via `broom` (Robinson et al., 2024).

<!-- In the next steps, we plan to expand these associations using generalized linear models (GLMs - McCullagh & Nelder, 1989) and generalized additive models (GAMs - Hastie & Tibshirani, 1990; Wood, 2017) to account for potential non-linear effects across the broader dataset. -->

## References

> R Core Team (2025). *R: A Language and Environment for Statistical Computing*. R Foundation for Statistical Computing, Vienna, Austria. <https://www.R-project.org/>.
>
> Wickham, H. (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York.
>
> Wickham, H., François, R., Henry, L., Müller, K., & Vaughan, D. (2023). *dplyr: A Grammar of Data Manipulation*. R package version 1.1.4.
>
> Wickham, H., Vaughan, D., & Girlich, M. (2024). *tidyr: Tidy Messy Data*. R package version 1.3.1.
>
> Wickham, H. (2023). *stringr: Simple, Consistent Wrappers for Common String Operations*. R package version 1.5.1.
>
> Robinson, D., Hayes, A., & Couch, S. (2024). *broom: Glance, Tidy, and Augment Regression Models*. R package version 1.0.6.
> 
> Muggeo, V. M. R. (2008). *segmented: an R Package to Fit Regression Models with Broken-Line Relationships*. R News, 8/1, 20-25.
>
> Garrett Grolemund, Hadley Wickham (2011). *Dates and Times Made Easy with lubridate*. Journal of Statistical Software, 40(3), 1-25.
<!-- > -->
<!-- > McCullagh, P., & Nelder, J. A. (1989). *Generalized Linear Models* (2nd ed.). Chapman & Hall. -->
<!-- > -->
<!-- > Hastie, T., & Tibshirani, R. (1990). *Generalized Additive Models*. Chapman & Hall. -->
<!-- > -->
<!-- > Wood, S. N. (2017). *Generalized Additive Models: An Introduction with R* (2nd ed.). CRC Press. -->