---
title: A standard multinomial GLMM with random correlated intercepts
author: Henrique Laureano (.github.io)
date: "*Last modification on* `r Sys.time()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r loading}
source('functions.R')
## install.packages('pacman')
pacman::p_load(Matrix, mvtnorm, mc2d, furrr, TMB, tictoc, tidyverse)
future::plan(multicore);TMB::openmp(11)

J <- 3e4
Z <- Matrix::bdiag(replicate(J, rep(1, 2), simplify=FALSE))
U <- matrix(0, nrow=J, ncol=2)
```

## easy data

```{r easy_data}
beta.easy <- c(1, 0.75) ## what makes it easy is the beta's choice
Sigma.0 <- matrix(c(1.0, 0.25, 0.25, 0.5), nrow=2, ncol=2)
## ch0l(Sigma.0) ## checking it positive-definiteness
set.seed(1237)
y.0 <- datasimu(
    J=J, beta=beta.easy, Sigma=Sigma.0
)
prop.table(colSums(y.0))
```

## model 0

> even with an "easy" dataset, i.e. with a good occurrence frequency in
> all multinomial classes, the model doesn't converge well - in this
> first moment I need to be able to get the "random" components in
> "sdreport()". Initial guesses at zero don't work well, so let's start
> with a GLM. If it is difficult now imagine later...

```{r glm}
glm.dll <- 'GLM';TMB::compile(paste0(glm.dll, '.cpp'))
checkDLL(glm.dll)
glm0.obj <- TMB::MakeADFun(data=list(Y=y.0),
                           parameters=list(beta1=0, beta2=0),
                           DLL=glm.dll, hessia=TRUE, silent=TRUE)
glm0.opt <- nlminb(glm0.obj$par,
                   glm0.obj$fn,
                   glm0.obj$gr)
glm0.opt
```

```{r m0,cache=TRUE,error=TRUE}
dll <- 'GLMM';TMB::compile(paste0(dll, '.cpp'))

checkDLL(dll)
obj0 <- TMB::MakeADFun(data=list(Y=y.0, Z=Z),
                       parameters=list(
                           beta1=glm0.opt$par['beta1'],
                           beta2=glm0.opt$par['beta2'],
                           logs2=c(log(0.1), log(0.1)),
                           rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                           U=U),
                       DLL=dll,
                       random='U', hessia=TRUE, silent=TRUE)
tictoc::tic()
opt0 <- nlminb(obj0$par,
               obj0$fn,
               obj0$gr)
tictoc::toc()
```

```{r m0_sdr,cache=TRUE,error=TRUE}
tictoc::tic()
sdr0 <- sdreport(obj0)
tictoc::toc()
```

```{r m0_results,cache=TRUE,error=TRUE}
opt0
summary(sdr0, select='fixed')
summary(sdr0, select='report')
obj0$report()
nrow(summary(sdr0, select='random'))
cbind(
    1:20,
    summary(sdr0, select='random')[1:20, ], 
    summary(sdr0, select='random')[30001:30020, ]
)
```

```{r v2m0,cache=TRUE,error=TRUE}
v2dll <- 'v2GLMM';TMB::compile(paste0(v2dll, '.cpp'))

checkDLL(v2dll)
v2obj0 <- TMB::MakeADFun(data=list(Y=y.0, Z=Z),
                         parameters=list(
                             beta1=glm0.opt$par['beta1'],
                             beta2=glm0.opt$par['beta2'],
                             logs2=c(log(0.1), log(0.1)),
                             rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                             U=U),
                         DLL=v2dll,
                         random='U', hessia=TRUE, silent=TRUE)
tictoc::tic()
v2opt0 <- nlminb(v2obj0$par,
                 v2obj0$fn,
                 v2obj0$gr)
tictoc::toc()
```

```{r v2m0_sdr,cache=TRUE,error=TRUE}
tictoc::tic()
v2sdr0 <- sdreport(v2obj0)
tictoc::toc()
```

```{r v2m0_results,cache=TRUE,error=TRUE}
v2opt0
summary(v2sdr0, select='fixed')
summary(v2sdr0, select='report')
v2obj0$report()
nrow(summary(v2sdr0, select='random'))
cbind(
    1:20,
    summary(v2sdr0, select='random')[1:20, ], 
    summary(v2sdr0, select='random')[30001:30020, ]
)
```

```{r fig.height=3.5}
dat <- tibble(
    value=summary(sdr0, select='random')[, 1],
    label=c(rep('u1', 3e4), rep('u2', 3e4))
)
dat%>%group_by(label)%>%summarise(mean=mean(value),
                                  min=min(value), 
                                  q025=quantile(value, probs=0.025),
                                  q50=quantile(value, probs=0.5),
                                  q975=quantile(value, probs=0.975),
                                  max=max(value))
ggplot(dat, aes(x=value))+geom_histogram()+facet_wrap(~label)
dat <- tibble(
    value=summary(v2sdr0, select='random')[, 1],
    label=c(rep('u1', 3e4), rep('u2', 3e4))
)
dat%>%group_by(label)%>%summarise(mean=mean(value),
                                  min=min(value), 
                                  q025=quantile(value, probs=0.025),
                                  q50=quantile(value, probs=0.5),
                                  q975=quantile(value, probs=0.975),
                                  max=max(value))
ggplot(dat, aes(x=value))+geom_histogram()+facet_wrap(~label)
```

## model 1

> let's generate 5 datasets from the same distributional scenario and
> see how the estimates vary

```{r five_data,cache=TRUE}
ys.5 <- future_datasimu(J=J, n=5, beta=beta.easy, Sigma=Sigma.0)
prop.table(colSums(ys.5[[1]]))
prop.table(colSums(ys.5[[2]]))
prop.table(colSums(ys.5[[3]]))
prop.table(colSums(ys.5[[4]]))
prop.table(colSums(ys.5[[5]]))
```

```{r five_guess}
guess5 <- getguess(dll=glm.dll,
                   y=ys.5, beta=c(beta1=beta.easy[1], beta2=beta.easy[2]))
```

```{r five_fit,cache=TRUE,error=TRUE}
tictoc::tic()
m1.1 <- GLMMfit(dll=v2dll, y=ys.5[[1]], Z=Z,
                pars=list(beta1=guess5[1, 'beta1'],
                          beta2=guess5[1, 'beta2'],
                          logs2=c(log(0.1), log(0.1)),
                          rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                          U=U))
tictoc::toc()
tictoc::tic()
m1.2 <- GLMMfit(dll=v2dll, y=ys.5[[2]], Z=Z,
                pars=list(beta1=guess5[2, 'beta1'],
                          beta2=guess5[2, 'beta2'],
                          logs2=c(log(0.1), log(0.1)),
                          rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                          U=U))
tictoc::toc()
tictoc::tic()
m1.3 <- GLMMfit(dll=v2dll, y=ys.5[[3]], Z=Z,
                pars=list(beta1=guess5[3, 'beta1'],
                          beta2=guess5[3, 'beta2'],
                          logs2=c(log(0.1), log(0.1)),
                          rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                          U=U))
tictoc::toc()
tictoc::tic()
m1.4 <- GLMMfit(dll=v2dll, y=ys.5[[4]], Z=Z,
                pars=list(beta1=guess5[4, 'beta1'],
                          beta2=guess5[4, 'beta2'],
                          logs2=c(log(0.1), log(0.1)),
                          rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                          U=U))
tictoc::toc()
tictoc::tic()
m1.5 <- GLMMfit(dll=v2dll, y=ys.5[[5]], Z=Z,
                pars=list(beta1=guess5[5, 'beta1'],
                          beta2=guess5[5, 'beta2'],
                          logs2=c(log(0.1), log(0.1)),
                          rhoZ=atanh( 0.05/(sqrt(0.1)*sqrt(0.1)) ),
                          U=U))
tictoc::toc()
```

```{r}
matrix(c(with(m1.1, c(par, objective, convergence)), 
         with(m1.4, c(par, objective, convergence)), 
         with(m1.5, c(par, objective, convergence))), nrow=3, byrow=TRUE,
       dimnames=list(c('1', '4', '5'), c(names(m1.1$par), 'fn', 'conv')))
```
