---
title: A standard multinomial GLMM
author: Henrique Laureano (.github.io)
date: "*Last modification on* `r Sys.time()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r loading}
source('functions.R')

## install.packages('pacman')
pacman::p_load(Matrix, mvtnorm, mc2d, ## rmultinomial()
               furrr, ## future + purrr
               TMB, tictoc, tidyverse ## ggplot2() + dplyr()
               )

future::plan(multicore);TMB::openmp(11)
```

## uniGLMM

> COMMON UNIVARIATE RANDOM INTERCEPT

> TAKE-HOME MESSAGE: PARALLELIZATION + ``PARAMETER_VECTOR()`` +
> ``dnorm()`` + MODELING THE ``log(sd)`` WORKS NICELY!

```{r uniGLMM,cache=TRUE,fig.height=3.5,fig.width=10}
J <- 250
cs <- 10
beta.mcie <- c(beta1=0.85, beta2=1.25)
sd <- sqrt(0.25)

set.seed(0858)
y <- datasimu.l1(J=J, cs=cs, beta=beta.mcie, sd=sd)
prop.table(colSums(y))

uniDLL <- 'uniGLMM';invisible(TMB::compile(paste0(uniDLL, '.cpp')))
## dyn.load(TMB::dynlib(uniDLL))
checkDLL(uniDLL)

Z <- modelZ(J=J, cs=cs)
u <- numeric(J)

obj <- TMB::MakeADFun(data=list(Y=y, Z=Z),
                      parameters=list(beta1=0,
                                      beta2=0,
                                      logsd=log(0.1),
                                      u=u),
                      DLL=uniDLL, ## ----------------------------------
                      random='u', hessian=TRUE, silent=TRUE)
(opt <- with(obj,
             nlminb(par, fn, gr)))
bind_cols(
    bind_rows(
        opt$par, c(beta.mcie, logsd=log(sd))
    ),
    label=c('model', 'data')
)

sdr <- sdreport(obj)
summary(sdr, select='fixed')
summary(sdr, select='report')
summary(summary(sdr, select='random'))

prof.beta1 <- TMB::tmbprofile(obj, name='beta1', trace=FALSE)
prof.beta2 <- TMB::tmbprofile(obj, name='beta2', trace=FALSE)
prof.logsd <- TMB::tmbprofile(obj, name='logsd', trace=FALSE)

par(mfrow=c(1, 3))
plot(prof.beta1);abline(v=beta.mcie['beta1'], lty=2, col='red')
plot(prof.beta2);abline(v=beta.mcie['beta2'], lty=2, col='red')
plot(prof.logsd);abline(v=log(sd), lty=2, col='red')
```

## uni2GLMM

> INDEPENDENT RANDOM INTERCEPTS

> TAKE-HOME MESSAGE: PARALLELIZATION + MULTIPLE ``PARAMETER_VECTOR()`` +
> MULTIPLE ``dnorm()`` + MODELING THE ``log(sd)`` WORKS NICELY!

```{r uni2GLMM,cache=TRUE,fig.height=6}
sd <- c(sqrt(0.25), sqrt(0.15))

set.seed(0859)
y <- datasimu.l1(J=J, cs=cs, beta=beta.mcie, sd=sd)
prop.table(colSums(y))

uni2DLL <- 'uni2GLMM';invisible(TMB::compile(paste0(uni2DLL, '.cpp')))
## dyn.load(TMB::dynlib(uni2DLL))
checkDLL(uni2DLL)

u1 <- u2 <- numeric(J)

obj <- TMB::MakeADFun(data=list(Y=y, Z=Z),
                      parameters=list(beta1=0,
                                      beta2=0,
                                      logsd1=log(0.1),
                                      logsd2=log(0.1),
                                      u1=u1,
                                      u2=u2),
                      DLL=uni2DLL, ## ----------------------------------
                      random=c('u1', 'u2'), hessian=TRUE, silent=TRUE)
(opt <- with(obj,
             nlminb(par, fn, gr)))
bind_cols(
    bind_rows(
        opt$par, c(beta.mcie, logsd1=log(sd[1]), logsd2=log(sd[2]))
    ),
    label=c('model', 'data')
)

sdr <- sdreport(obj)
summary(sdr, select='fixed')
summary(sdr, select='report')
summary(summary(sdr, select='random'))

prof.beta1 <- TMB::tmbprofile(obj, name='beta1', trace=FALSE)
prof.beta2 <- TMB::tmbprofile(obj, name='beta2', trace=FALSE)
prof.logsd1 <- TMB::tmbprofile(obj, name='logsd1', trace=FALSE)
prof.logsd2 <- TMB::tmbprofile(obj, name='logsd2', trace=FALSE)

par(mfrow=c(2, 2), mar=c(4, 4, 2, 2))
plot(prof.beta1);abline(v=beta.mcie['beta1'], lty=2, col='red')
plot(prof.beta2);abline(v=beta.mcie['beta2'], lty=2, col='red')
plot(prof.logsd1);abline(v=log(sd[1]), lty=2, col='red')
plot(prof.logsd2);abline(v=log(sd[2]), lty=2, col='red')
```

## uni3GLMM

> INDEPENDENT RANDOM INTERCEPTS

> TAKE-HOME MESSAGE: PARALLELIZATION + ``PARAMETER_MATRIX()`` + MULTIPLE
> ``dnorm()`` + MODELING THE ``log(sd)`` WORKS NICELY!

```{r uni3GLMM,cache=TRUE,fig.height=6}
uni3DLL <- 'uni3GLMM';invisible(TMB::compile(paste0(uni3DLL, '.cpp')))
## dyn.load(TMB::dynlib(uni3DLL))
checkDLL(uni3DLL)

U <- matrix(0, nrow=J, ncol=2)

obj <- TMB::MakeADFun(data=list(Y=y, Z=Z),
                      parameters=list(beta1=0,
                                      beta2=0,
                                      logsd1=log(0.1),
                                      logsd2=log(0.1),
                                      U=U),
                      DLL=uni3DLL, ## ----------------------------------
                      random='U', hessian=TRUE, silent=TRUE)
(opt <- with(obj,
             nlminb(par, fn, gr)))
bind_cols(
    bind_rows(
        opt$par, c(beta.mcie, logsd1=log(sd[1]), logsd2=log(sd[2]))
    ),
    label=c('model', 'data')
)

sdr <- sdreport(obj)
summary(sdr, select='fixed')
summary(sdr, select='report')
summary(sdr, select='random')[251, ]
summary(summary(sdr, select='random'))

prof.beta1 <- TMB::tmbprofile(obj, name='beta1', trace=FALSE)
prof.beta2 <- TMB::tmbprofile(obj, name='beta2', trace=FALSE)
prof.logsd1 <- TMB::tmbprofile(obj, name='logsd1', trace=FALSE)
prof.logsd2 <- TMB::tmbprofile(obj, name='logsd2', trace=FALSE)

par(mfrow=c(2, 2), mar=c(4, 4, 2, 2))
plot(prof.beta1);abline(v=beta.mcie['beta1'], lty=2, col='red')
plot(prof.beta2);abline(v=beta.mcie['beta2'], lty=2, col='red')
plot(prof.logsd1);abline(v=log(sd[1]), lty=2, col='red')
plot(prof.logsd2);abline(v=log(sd[2]), lty=2, col='red')
```

## uni4GLMM

> INDEPENDENT RANDOM INTERCEPTS

> TAKE-HOME MESSAGE: PARALLELIZATION + ``PARAMETER_MATRIX()`` +
> ``MVNORM()`` + MODELING THE ``log(sd)`` WORKS NICELY!

```{r uni4GLMM,cache=TRUE,fig.height=6}
uni4DLL <- 'uni4GLMM';invisible(TMB::compile(paste0(uni4DLL, '.cpp')))
dyn.load(TMB::dynlib(uni4DLL))
checkDLL(uni4DLL)

obj <- TMB::MakeADFun(data=list(Y=y, Z=Z),
                      parameters=list(beta1=0,
                                      beta2=0,
                                      logsd1=log(0.1),
                                      logsd2=log(0.1),
                                      U=U),
                      DLL=uni4DLL, ## ----------------------------------
                      random='U', hessian=TRUE, silent=TRUE)
(opt <- with(obj,
             nlminb(par, fn, gr)))
bind_cols(
    bind_rows(
        opt$par, c(beta.mcie, logsd1=log(sd[1]), logsd2=log(sd[2]))
    ),
    label=c('model', 'data')
)
obj$report()
sdr <- sdreport(obj)
summary(sdr, select='fixed')
summary(sdr, select='report')
summary(sdr, select='random')[251, ]
summary(summary(sdr, select='random'))

prof.beta1 <- TMB::tmbprofile(obj, name='beta1', trace=FALSE)
prof.beta2 <- TMB::tmbprofile(obj, name='beta2', trace=FALSE)
prof.logsd1 <- TMB::tmbprofile(obj, name='logsd1', trace=FALSE)
prof.logsd2 <- TMB::tmbprofile(obj, name='logsd2', trace=FALSE)

par(mfrow=c(2, 2), mar=c(4, 4, 2, 2))
plot(prof.beta1);abline(v=beta.mcie['beta1'], lty=2, col='red')
plot(prof.beta2);abline(v=beta.mcie['beta2'], lty=2, col='red')
plot(prof.logsd1);abline(v=log(sd[1]), lty=2, col='red')
plot(prof.logsd2);abline(v=log(sd[2]), lty=2, col='red')
```

## uni5GLMM

> INDEPENDENT RANDOM INTERCEPTS

> TAKE-HOME MESSAGE: PARALLELIZATION + ``PARAMETER_MATRIX()`` +
> ``MVNORM()`` + MODELING THE ``log(s2)`` DOESN'T WORK!
> PROBLEM/DIFFERENCE? WE'RE MODELING THE ``log(s2)`` NOT THE ``log(sd)``

```{r uni5GLMM,cache=TRUE,error=TRUE}
uni5DLL <- 'uni5GLMM';invisible(TMB::compile(paste0(uni5DLL, '.cpp')))
dyn.load(TMB::dynlib(uni5DLL))
checkDLL(uni5DLL)

obj <- TMB::MakeADFun(data=list(Y=y, Z=Z),
                      parameters=list(beta1=beta.mcie['beta1'],
                                      beta2=beta.mcie['beta2'],
                                      logs2_1=log(0.25),
                                      logs2_2=log(0.15),
                                      U=U),
                      DLL=uni5DLL, ## ----------------------------------
                      random='U', hessian=TRUE, silent=TRUE)
(opt <- with(obj,
             nlminb(par, fn, gr)))

sdr <- sdreport(obj)
summary(sdr, select='fixed')
summary(sdr, select='report')
summary(summary(sdr, select='random'))
```
