---
title: "Modeling the cumulative incidence function\\newline
        of clustered competing risk data: \\newline
        computational and numerical aspects\\newline
        of a multinomial GLMM approach"
subtitle: ""
short-title: ""
author: "\\includegraphics[width=.175\\textwidth]{logo/leg.pdf}
         \\hspace{.5\\textwidth}
         \\includegraphics[width=.25\\textwidth]{logo/ufpr.png}
         \\vspace{.75cm}\\newline
         Henrique Ap. Laureano\\hspace{.25\\textwidth}
         Wagner H. Bonat\\newline
         \\url{http://leg.ufpr.br/~henrique} $\\mid$
         \\url{http://leg.ufpr.br/~wagner}"
short-author: "leg.ufpr.br/~henrique"
# email: "laureano\\@ufpr.br"
# date: "*Last modification on* `r Sys.time()`"
date: "December/2020"
short-date: "leg.ufpr.br/~henrique"
department: ""
institute: ""
# short-institute: "LEG"
section-titles: true
safe-columns: true # enables special latex macros for columns
output:
  beamer_presentation:
    template: template.tex
---

```{r, include=FALSE}
library(knitr)

knit_hooks$set(rmdsize = function(before, options, envir) {
    if (before) "\\footnotesize"
    else "\\normalsize"
})

opts_chunk$set(rmdsize = TRUE,
               warning = FALSE,
               cache = TRUE,
               cache_path = "min15_cache/",
               fig.path = "min15_figures/")
```

### 

\begin{minipage}{.525\textwidth}
 \begin{shaded}
  Context: clustered competing risk data
 \end{shaded}
\end{minipage}

Idea: causes competing by the occurence of an event such the

\begin{minipage}{.425\textwidth}
 \begin{block}{confiability analysis}
  failure of an industrial or electronic component
 \end{block}
\end{minipage}
\hspace{0.5cm}
\begin{minipage}{.425\textwidth}
 \begin{block}{survival analysis}
  failure or progress of a patient or some biological process
 \end{block}
\end{minipage}

\vspace{0.5cm}
\pause

A typical data set consists of

\begin{tabular}{lc*{3}{>{\columncolor[gray]{0.8}}c}cr}
 Group & ID & Cause 1 & Cause 2 & Censorship & Time & Feature\\
 \hline
 1     & 1  & 1       & 0       & 0          & 10   & A\\
 1     & 2  & 0       & 0       & 1          &  8   & A\\
 2     & 1  & 0       & 0       & 1          &  7   & B\\
 2     & 2  & 0       & 1       & 0          &  5   & A\\
\end{tabular}

###

\begin{minipage}{.275\textwidth}
 \begin{shaded}
  Survival data designs
 \end{shaded}
\end{minipage}

\begin{center}
 \scalebox{0.725}{
  \begin{tikzpicture}
   \begin{pgfonlayer}{nodelayer}
    \node [style=circle] (2) at (-17.5, 5.5) {0};
    \node [style=circle] (3) at (-13.5, 5.5) {1};
    \node [style=circle] (7) at (-7.5, 5.5) {1};
    \node [style=circle] (9) at (-7.5, 4.5) {2};
    \node [style=none] (11) at (-7.5, 3.5) {$\vdots$};
    \node [style=circle] (12) at (-7.5, 2.5) {$m$};
    \node [style=circle] (14) at (-12.25, 4) {0};
    \node [style=circle] (22) at (-6, 3.5) {0};
    \node [style=circle] (23) at (-2, 3.5) {2};
    \node [style=circle] (25) at (-4, 5.5) {1};
    \node [style=none] (27) at (-17, 5.5) {};
    \node [style=none] (28) at (-14, 5.5) {};
    \node [style=none] (29) at (-11.75, 4.25) {};
    \node [style=none] (30) at (-8, 5.5) {};
    \node [style=none] (31) at (-8, 4.5) {};
    \node [style=none] (32) at (-8, 2.5) {};
    \node [style=none] (33) at (-5.5, 3.5) {};
    \node [style=none] (34) at (-4.5, 5.25) {};
    \node [style=none] (35) at (-2.5, 3.5) {};
    \node [style=none] (36) at (-11.75, 4) {};
    \node [style=none] (37) at (-11.75, 3.75) {};
    \node [style=none] (38) at (-5.5, 3.75) {};
    \node [style=none] (39) at (-2.5, 3.75) {};
    \node [style=none] (40) at (-3.5, 5.25) {};
    \node [style=none] (44) at (-15.5, 6.75) {{\Large
                                               Failure time process}};
    \node [style=none] (45) at (-10, 6.75) {{\Large
                                             Competing risk process}};
    \node [style=none] (47) at (-4, 6.75) {{\Large Multistate process}};
   \end{pgfonlayer}
   \begin{pgfonlayer}{edgelayer}
	\draw [style=1side] (27.center) to (28.center);
	\draw [style=1side] (37.center) to (32.center);
	\draw [style=1side] (40.center) to (39.center);
	\draw [style=1side] (33.center) to (35.center);
	\draw [style=2side] (38.center) to (34.center);
	\draw [style=1side] (36.center) to (31.center);
	\draw [style=1side] (29.center) to (30.center);
   \end{pgfonlayer}
  \end{tikzpicture}
 }
\end{center}

\pause
\begin{minipage}{.375\textwidth}
 \begin{shaded}
  Survival modeling framework
 \end{shaded}
\end{minipage}

We have to choose which \textcolor{mykindofblue}{scale} we model the
\textbf{survival experience}.\newline
Usually, is the
\[
 \text{\textcolor{mykindofblue}{hazard} (failure rate)
       \textcolor{mykindofblue}{scale}}: \quad
 \lambda(t \mid x) = \lambda_{0}(t) \times c(x, \beta)
\]

###

\begin{minipage}{.425\textwidth}
 \begin{shaded}
  In the competing risk setting \(\dots\)
 \end{shaded}
\end{minipage}

a more attractive possibility is to work on the
\textcolor{mykindofblue}{probability scale},
focusing on the cause-specific

```{r cif, echo=FALSE, fig.height=2, fig.width=4}
pacman::p_load(tidyverse, ## you need the pacman's package
               ggplot2)
t <- seq(30, 79.5, by = 0.5)
delta <- 80
u <- c(0, 0)
eta <- c(0, 0)
beta <- c(-2, -1.5)
gamma <- c(1.2, 1)
w <- c(3, 5)
risk1 <- exp(beta[1] + u[1])
risk2 <- exp(beta[2] + u[2])
level <- 1 +  risk1 + risk2
risklevel1 <- risk1/level
risklevel2 <- risk2/level
traj1 <- pnorm(w[1] * atanh(2 * t/delta - 1) - gamma[1] - eta[1])
traj2 <- pnorm(w[2] * atanh(2 * t/delta - 1) - gamma[2] - eta[2])
cif1 <- risklevel1 * traj1
cif2 <- risklevel2 * traj2
df <- as.data.frame(cbind(t, cif1, cif2))
df <- df %>% pivot_longer(cols = cif1:cif2,
                          names_to = "cif", values_to = "value")
ggplot(df, aes(t, value, group = cif, color = cif)) +
    geom_line(size = 1.5) +
    labs(x = "Time", y = "Incidence",
         title = 'Cumulative Incidence Function (CIF)') +
    scale_color_manual(values = c("#ff7f00", "#0080ff"),
                       labels = c("Cause 1", "Cause 2")) +
    theme(legend.title = element_blank(),
          legend.text = element_text(size = 8),
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          plot.title = element_text(
              size = 9,
              margin = unit(c(t = 0, r = 0, b = 2, l = 0), 'mm')), 
          axis.title.x = element_text(
              size = 8,
              margin = unit(c(t = 2, r = 0, b = 0, l = 0), "mm")),
          axis.title.y = element_text(
              size = 8,
              margin = unit(c(t = 0, r = 2, b = 0, l = 0), "mm")))
```

i.e.
\[
 \text{CIF} = \mathbb{P}[~\text{failure time} \leq t,
                         ~\text{a given cause} \mid \text{features}~]
\]

###

\begin{minipage}{.65\textwidth}
 \begin{shaded} Main focus application: cancer incidence in twins
 \end{shaded}
\end{minipage}

\vspace{0.5cm}

\begin{columns}
 \begin{column}{.55\textwidth}
  \centering
   \includegraphics[width=\textwidth]{figures/twins.jpg}
 \end{column}
\begin{column}{.45\textwidth}
 Clustered competing risks data\newline
 \hspace*{.5cm}\rotatebox[origin=c]{180}{$\Lsh$}
 Clusters? Families\newline
 \hspace*{1.35cm}\rotatebox[origin=c]{180}{$\Lsh$}Family studies\newline
 \hspace*{2.1cm}\rotatebox[origin=c]{180}{$\Lsh$}
 \textcolor{mykindofblue}{Twins data}
 \end{column}
\end{columns}

\pause
+ Taking into account the within-family dependence may reflect both
  disease \textbf{heritability} and the impact of shared
  \textbf{environmental effects}

+ A complication is that we have little information to track that
  \textbf{dependence} since each 'family' consists of only a pair of
  twins

###

\begin{minipage}{.15\textwidth}
 \begin{shaded}
  Challenges
 \end{shaded}
\end{minipage}

Besides the small size groups, the data is very simple \(\dots\)

+ we just know if the event occured (1 or 0) and the time

  + with this, we have to be able to construct the\newline
    \textcolor{mykindofblue}{cumulative incidence curves}

+ and we have to accommodate the
  \textcolor{mykindofblue}{within-family dependency}

  + that can happen in different ways and with different intensities

\noindent{\color{notthefruitorange}\rule{\linewidth}{0.25mm}}

to accomplish all this a powerful modeling framework is made necessary

\(\dots\) with this,\newline
computational and numerical challenges has also to be overcome

###

\begin{minipage}{.15\textwidth}
 \begin{shaded}
  Thank you
 \end{shaded}
\end{minipage}

\vspace{0.5cm}

\begin{columns}
 \begin{column}{.33\textwidth}
  \centering
   \includegraphics[width=.9\textwidth]{logo/ppgmne.png}
 \end{column}
 \begin{column}{.33\textwidth}
  \centering
   \includegraphics[width=.9\textwidth]{logo/capes.jpg}
 \end{column}
 \begin{column}{.33\textwidth}
  \centering
   \includegraphics[width=.9\textwidth]{logo/ufpr.png}
 \end{column}
\end{columns}

\vspace{0.5cm}
\noindent{\color{notthefruitorange}\rule{\linewidth}{0.25mm}}

\url{http://leg.ufpr.br/~henrique}

[\@hap_laureano](https://twitter.com/hap_laureano)
