---
title: "Modeling the cumulative incidence function\\newline
        of clustered competing risk data"
subtitle: ""
short-title: ""
author: "\\includegraphics[width=.175\\textwidth]{logo/leg.pdf}
         \\hspace{.5\\textwidth}
         \\includegraphics[width=.25\\textwidth]{logo/ufpr.png}
         \\vspace{.75cm}\\newline
         Henrique Ap. Laureano\\newline
         \\url{http://leg.ufpr.br/~henrique}"
short-author: "leg.ufpr.br/~henrique"
# email: "laureano\\@ufpr.br"
# date: "*Last modification on* `r Sys.time()`"
date: "December/2020"
short-date: "leg.ufpr.br/~henrique"
department: ""
institute: ""
# short-institute: "LEG"
section-titles: true
safe-columns: true # enables special latex macros for columns
output:
  beamer_presentation:
    template: template.tex
---

```{r, include=FALSE}
library(knitr)

knit_hooks$set(rmdsize = function(before, options, envir) {
    if (before) "\\footnotesize"
    else "\\normalsize"
})

opts_chunk$set(rmdsize = TRUE,
               warning = FALSE,
               cache = TRUE,
               cache_path = "min3_cache/",
               fig.path = "min3_figures/")
```

### 

\begin{minipage}{.4\textwidth}
 \begin{shaded}
  Clustered competing risk data
 \end{shaded}
\end{minipage}

Key terms:

+ \textcolor{mykindofblue}{Clustered}:
  groups with a dependence structure (e.g. families);

+ Causes \textcolor{mykindofblue}{competing} by *something*.

Something?

+ **Failure** of an industrial or electronic component;

+ **Occurence** or **cure** of a disease;

+ **Progress** of a patient clinic state.

Independent of the application, always the same framework

\begin{tabular}{lc*{3}{>{\columncolor[gray]{0.8}}c}r}
 Cluster & ID & Cause 1 & Cause 2 & Censorship & Time\\
 \hline
 1     & 1  & Yes     & No       & No        & 10\\
 1     & 2  & No      & No       & Yes       &  8\\
 2     & 1  & No      & No       & Yes       &  7\\
 2     & 2  & No      & Yes      & No        &  5\\
\end{tabular}

###

\begin{minipage}{.175\textwidth}
 \begin{shaded}
  What we do?
 \end{shaded}
\end{minipage}

\vspace{0.5cm}

We model the \textcolor{mykindofblue}{probability} of each **competing
cause** along the time and taking into account the possible
\textcolor{mykindofblue}{within-cluster dependence}

\(\dots\) all this in terms of a

\vspace{0.5cm}
```{r cif, echo=FALSE, fig.height=2, fig.width=4}
pacman::p_load(tidyverse, ## you need the pacman's package
               ggplot2)
t <- seq(30, 79.5, by = 0.5)
delta <- 80
u <- c(0, 0)
eta <- c(0, 0)
beta <- c(-2, -1.5)
gamma <- c(1.2, 1)
w <- c(3, 5)
risk1 <- exp(beta[1] + u[1])
risk2 <- exp(beta[2] + u[2])
level <- 1 +  risk1 + risk2
risklevel1 <- risk1/level
risklevel2 <- risk2/level
traj1 <- pnorm(w[1] * atanh(2 * t/delta - 1) - gamma[1] - eta[1])
traj2 <- pnorm(w[2] * atanh(2 * t/delta - 1) - gamma[2] - eta[2])
cif1 <- risklevel1 * traj1
cif2 <- risklevel2 * traj2
df <- as.data.frame(cbind(t, cif1, cif2))
df <- df %>% pivot_longer(cols = cif1:cif2,
                          names_to = "cif", values_to = "value")
ggplot(df, aes(t, value, group = cif, color = cif)) +
    geom_line(size = 1.5) +
    labs(x = "Time", y = "Incidence",
         title = 'Cumulative Incidence Function') +
    scale_color_manual(values = c("#ff7f00", "#0080ff"),
                       labels = c("Cause 1", "Cause 2")) +
    theme(legend.title = element_blank(),
          legend.text = element_text(size = 8),
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          plot.title = element_text(
              size = 9,
              margin = unit(c(t = 0, r = 0, b = 2, l = 0), 'mm')), 
          axis.title.x = element_text(
              size = 8,
              margin = unit(c(t = 2, r = 0, b = 0, l = 0), "mm")),
          axis.title.y = element_text(
              size = 8,
              margin = unit(c(t = 0, r = 2, b = 0, l = 0), "mm")))
```

###

\begin{minipage}{.65\textwidth}
 \begin{shaded} Main focus application: cancer incidence in twins
 \end{shaded}
\end{minipage}

\vspace{0.5cm}

\begin{columns}
 \begin{column}{.55\textwidth}
  \centering
   \includegraphics[width=\textwidth]{figures/twins.jpg}
 \end{column}
\begin{column}{.45\textwidth}
 Clustered competing risks data\newline
 \hspace*{.5cm}\rotatebox[origin=c]{180}{$\Lsh$}
 Clusters? Families\newline
 \hspace*{1.35cm}\rotatebox[origin=c]{180}{$\Lsh$}Family studies\newline
 \hspace*{2.1cm}\rotatebox[origin=c]{180}{$\Lsh$}
 \textcolor{mykindofblue}{Twins data}
 \end{column}
\end{columns}

\[
 \text{Family studies}~\Rightarrow~
 \text{\color{mykindofblue}{within-family dependence}}
\]
That may reflect

+ Disease \textbf{heritability};
  
+ The impact of shared \textbf{environmental effects};
  
  + **Parental effects**:\newline
    continuity of the phenotype across generations.

###

\begin{minipage}{.15\textwidth}
 \begin{shaded}
  Challenges
 \end{shaded}
\end{minipage}

\vspace{0.5cm}

+ The data is very simple,\newline
  we just know the outcome (**yes** or **no**);

+ With this, we have to be able to construct the\newline
  \textcolor{mykindofblue}{cumulative incidence curves};

+ And we have to accommodate the
  \textcolor{mykindofblue}{within-family dependence} properly,\newline
  that can happen in different manners;

+ \(\dots\)

###

\begin{minipage}{.2\textwidth}
 \begin{shaded}
  Joint work with
 \end{shaded}
\end{minipage}

\vspace{0.5cm}

\begin{columns}
 \begin{column}{.5\textwidth}
   Wagner H. Bonat\newline
   \url{http://leg.ufpr.br/~wagner}
 \end{column}
 \begin{column}{.5\textwidth}
  Paulo Justiniano Ribeiro Jr.\newline
  \url{http://leg.ufpr.br/~paulojus}
 \end{column}
\end{columns}

\vspace{0.5cm}

\begin{minipage}{.15\textwidth}
 \begin{shaded}
  Thank you
 \end{shaded}
\end{minipage}

\begin{columns}
 \begin{column}{.33\textwidth}
  \centering
   \includegraphics[width=.9\textwidth]{logo/ppgmne.png}
 \end{column}
 \begin{column}{.33\textwidth}
  \centering
   \includegraphics[width=.9\textwidth]{logo/capes.jpg}
 \end{column}
 \begin{column}{.33\textwidth}
  \centering
   \includegraphics[width=.9\textwidth]{logo/ufpr.png}
 \end{column}
\end{columns}

\vspace{0.5cm}

\url{http://leg.ufpr.br/~henrique}

[\@hap_laureano](https://twitter.com/hap_laureano)
