---
title: "Modeling the cumulative incidence function\\newline
        of clustered competing risk data: \\newline
        a multinomial GLMM approach"
subtitle: ""
short-title: ""
author: "\\includegraphics[width=.175\\textwidth]{logo/leg.pdf}
         \\hspace{.5\\textwidth}
         \\includegraphics[width=.25\\textwidth]{logo/ufpr.png}
         \\vspace{.75cm}\\newline
         Henrique Laureano\\hspace{.25\\textwidth} Wagner Bonat\\newline
         \\url{http://leg.ufpr.br/~henrique} $\\mid$
         \\url{http://leg.ufpr.br/~wagner}"
short-author: "leg.ufpr.br/~henrique"
# email: "laureano\\@ufpr.br"
# date: "*Last modification on* `r Sys.time()`"
short-date: "leg.ufpr.br/~henrique"
department: ""
institute: ""
# short-institute: "LEG"
section-titles: true
safe-columns: true # enables special latex macros for columns
output:
  legtheme::beamer_leg
---

```{r, include=FALSE}
library(knitr)

knit_hooks$set(rmdsize = function(before, options, envir) {
    if (before) "\\footnotesize"
    else "\\normalsize"
})

opts_chunk$set(rmdsize = TRUE,
               warning = FALSE,
               cache = TRUE,
               cache_path = "slides_cache/",
               fig.path = "slides_figures/")
```

###

\begin{minipage}{.775\textwidth}
 \begin{shaded}
  Cause-specific \textbf{cumulative incidence function} (CIF),
  \(F(\cdot)\)
 \end{shaded}
\end{minipage}

\[
\begin{aligned}
  \text{for a type}~{\color{beamer@UIUCorange}{k}}~\text{failure},
  \quad
  F_{\color{beamer@UIUCorange}{k}}(t\mid\{X, R\})
  &=\mathbb{P}[T\leq t,
               K = {\color{beamer@UIUCorange}{k}}\mid\{X, R\}]\\
  &=\underbrace{
     \pi_{\color{beamer@UIUCorange}{k}}(\{X, R\})}_{
     \substack{\text{risk level}}}\times
    \underbrace{
     \Phi[t\mid\{X, R_{\color{beamer@UIUCorange}{k}}\}]}_{
     \substack{\text{failure time}\\\text{trajectory}}}.
\end{aligned}
\]
As proposed by \hyperlink{SCHEIKE}{Cederkvist et al., 2019}.

\pause
```{r cif, echo=FALSE, fig.height=2, fig.width=4}
pacman::p_load(tidyverse, ## you need the pacman's package
               ggplot2)
t <- seq(30, 79.5, by = 0.5)
delta <- 80
u <- c(0, 0)
eta <- c(0, 0)
beta <- c(-2, -1.5)
gamma <- c(1.2, 1)
w <- c(3, 5)
risk1 <- exp(beta[1] + u[1])
risk2 <- exp(beta[2] + u[2])
level <- 1 +  risk1 + risk2
risklevel1 <- risk1/level
risklevel2 <- risk2/level
traj1 <- pnorm(w[1] * atanh(2 * t/delta - 1) - gamma[1] - eta[1])
traj2 <- pnorm(w[2] * atanh(2 * t/delta - 1) - gamma[2] - eta[2])
cif1 <- risklevel1 * traj1
cif2 <- risklevel2 * traj2
df <- as.data.frame(cbind(t, cif1, cif2))
df <- df %>% pivot_longer(cols = cif1:cif2,
                          names_to = "cif", values_to = "value")
ggplot(df, aes(t, value, group = cif, color = cif)) +
    geom_line(size = 1.5) +
    labs(x = "Time", y = "Incidence") +
    scale_color_manual(values = c("#ff7f00", "#0080ff"),
                       labels = c("Cause 1", "Cause 2")) +
    theme(legend.title = element_blank(),
          legend.text = element_text(size = 8),
          axis.text.x = element_text(size = 7),
          axis.text.y = element_text(size = 7),
          axis.title.x = element_text(
              size = 8,
              margin = unit(c(t = 2, r = 0, b = 0, l = 0), "mm")),
          axis.title.y = element_text(
              size = 8,
              margin = unit(c(t = 0, r = 2, b = 0, l = 0), "mm")))
```

###

\begin{minipage}{.275\textwidth}
 \begin{shaded}
  Survival data designs
 \end{shaded}
\end{minipage}

###

+ We work on the probability scale, modeling the

\begin{block}{Cause-specific \textbf{cumulative incidence function}
(CIF)}
\[
\begin{aligned}
  \text{for a type}~{\color{beamer@UIUCorange}{k}}~\text{failure},
  \quad
  F_{\color{beamer@UIUCorange}{k}}(t \mid X) &=
  \mathbb{P}[T \leq t,
  {\color{beamer@UIUCorange}{K}} =
  {\color{beamer@UIUCorange}{k}} \mid X]\\
  &= \int_{0}^{t} f_{\color{beamer@UIUCorange}{k}}(a \mid X) da,
  \quad t > 0,
\end{aligned}
\]
where \(f_{\color{beamer@UIUCorange}{k}}(t \mid X) =
        \lambda_{\color{beamer@UIUCorange}{k}}(t \mid X) \times
        S(t \mid X)\) is the (sub)density for the time to a type
        \(\color{beamer@UIUCorange}{k}\) failure
\end{block}

+ We go beyond the independence supposition, having clusters of subjects
  with a possible correlation

+ Our latent structure to capture this possible correlation has\newline
  audacity, allowing for \textcolor{beamer@UIUCblue}{within-cluster
  dependence} of both\newline risk and timing

###

Considering a competing risks setup of three competing causes\dots

\pause
\dots we want to model the \textcolor{beamer@UIUCblue}{CIF} in the
following specific manner
\[
\begin{aligned}
  F_{\color{beamer@UIUCorange}{k}}(t \mid X, U) &=
  \mathbb{P}[T \leq t, {\color{beamer@UIUCorange}{K}} =
                       {\color{beamer@UIUCorange}{k}} \mid X, U]\\
  &= \underbrace{
    \pi_{\color{beamer@UIUCorange}{k}}(X, u_{1}, u_{2})}_{
    \substack{\text{cluster-specific}\\\text{risk level}}}
  \times
  \underbrace{
    \Phi[w_{\color{beamer@UIUCorange}{k}} g(t) -
         X^{\top}\gamma_{\color{beamer@UIUCorange}{k}} -
         \eta_{\color{beamer@UIUCorange}{k}}]}_{
         \substack{\text{cluster-specific}\\
                   \text{failure time trajectory}}},
               \quad {\color{beamer@UIUCorange}{k}} = 1, 2.
\end{aligned}
\]
i.e., a product of two functions

+ the first models the \textcolor{beamer@UIUCblue}{absolute risk} of a
  competing cause in the presence of another

+ the second models the \textcolor{beamer@UIUCblue}{failure time
  distribution},

both varying between clusters

\noindent{\color{beamer@UIUCorange}\rule{\linewidth}{0.25mm}}
\pause

+ The third competing cause is left to accommodate\newline
  the \textcolor{beamer@UIUCblue}{censorship}

+ Which probability distribution fits this setup?
  \textcolor{beamer@UIUCblue}{Multinomial}

###

\texttt{Scheike} works out this idea via a pairwise composite likelihood
approach,

\small
```
@article{Scheike,
  author  = {L. Cederkvist and K. K. Holst and K. K. Andersen and
             T. H. Scheike},
  title   = {Modeling the cumulative incidence function of
             multivariate competing risks data allowing for
             within-cluster dependence of risk and timing},
  year    = {2019},
  journal = {Biostatistics},
  volume  = {20},
  number  = {2},
  pages   = {199--217},
}
```
\vspace{-0.5cm}
\normalsize
\noindent{\color{beamer@UIUCorange}\rule{\linewidth}{0.25mm}}

And we? We try a hierarchical approach via a GLMM

###

\begin{minipage}{.35\textwidth}
 \begin{shaded} Our approach: what we do
 \end{shaded}
\end{minipage}
\pause

For an observed response \(y_{ijt}\) of a subject \(i\), with cluster
\(j\), in the time \(t\), we have
\[
\begin{aligned}
  y_{ijt} \mid \{u_{1j}, u_{2j}, \eta_{1j}, \eta_{2j}\} &\sim
  \text{Multinomial}(p_{1ijt}, p_{2ijt}, p_{3ijt}),\\
  \begin{bmatrix} u_{1j}\\u_{2j}\\\eta_{1j}\\\eta_{2j} \end{bmatrix}
  &\sim
  \mathcal{N}
  \left[\begin{bmatrix} 0\\0\\0\\0\end{bmatrix},
    \begin{bmatrix}
      \sigma_{u_{1}}^{2}&\varrho_{u_{1}, u_{2}}&
      \varrho_{u_{1}, \eta_{1}}&\varrho_{u_{1}, \eta_{2}}\\
      &\sigma_{u_{2}}^{2}&
      \varrho_{u_{2}, \eta_{1}}&\varrho_{u_{2}, \eta_{2}}\\
      &&\sigma_{\eta_{1}}^{2}&\varrho_{\eta_{1}, \eta_{2}}\\
      &&&\sigma_{\eta_{2}}^{2}
    \end{bmatrix}
  \right]
\end{aligned}.
\]
\pause
This model structure is connected via a \textcolor{beamer@UIUCblue}{link
function}, i.e., a functional form for the multinomial probabilities.

Here we use the CIF derivative w.r.t. time \(t\),

\begin{minipage}{0.8\textwidth}
\[
\begin{aligned}
  p_{{\color{beamer@UIUCorange}{k}}ijt} &=
  \frac{\exp\{
    \bm{x}_{{\color{beamer@UIUCorange}{k}}ij}
    \bm{\beta}_{{\color{beamer@UIUCorange}{k}}i} +
    u_{{\color{beamer@UIUCorange}{k}}j}\}}{
    1 + \sum_{{\color{beamer@UIUCorange}{m}}=1}^{
      {\color{beamer@UIUCorange}{K}}-1
    }
    \exp\{
    \bm{x}_{{\color{beamer@UIUCorange}{m}}ij}
    \bm{\beta}_{{\color{beamer@UIUCorange}{m}}i} +
    u_{{\color{beamer@UIUCorange}{m}}j}\}}
  \hspace{1.85cm}\{{\color{beamer@UIUCorange}{k}} = 1, 2\}\\
  &\times
  w_{\color{beamer@UIUCorange}{k}}\frac{\delta}{2\delta t - 2t^{2}}
  \phi[
  w_{\color{beamer@UIUCorange}{k}}
  \text{arctanh}\left(\frac{t-\delta/2}{\delta/2}\right) -
  \bm{x}_{{\color{beamer@UIUCorange}{k}}ij}
  \bm{\gamma}_{{\color{beamer@UIUCorange}{k}}i} -
  \eta_{{\color{beamer@UIUCorange}{k}}j}
  ].
\end{aligned}
\]
\end{minipage}

###

\begin{minipage}{.35\textwidth}
 \begin{shaded} Our approach: how we do
 \end{shaded}
\end{minipage}
\pause

\begin{center}
\textcolor{beamer@UIUCblue}{TMB: Template Model Builder}
\end{center}

Quickly implement complex random effect models through simple
\texttt{C++} templates. The \texttt{R} package combines

+ \texttt{CppAD}: \texttt{C++} automatic differentiation

+ \texttt{Eigen}: templated matrix-vector library

+ \texttt{CHOLMOD}: sparse matrix routines available from \texttt{R}

to obtain an efficient implementation of the applied Laplace
approximation with exact derivatives.

Also, key features are

+ automatic sparseness detection

+ parallelism through \texttt{BLAS}

+ parallel user templates.

###

\begin{minipage}{.6\textwidth}
 \begin{shaded} Main application: cancer incidence in twins
 \end{shaded}
\end{minipage}
\vspace{0.75cm}
\begin{columns}
\begin{column}{.55\linewidth}
\centering
\includegraphics[width=\linewidth]{prints/twins.jpg}
\end{column}
\begin{column}{.45\linewidth}
Clustered competing risks data\newline
\hspace*{.5cm}\rotatebox[origin=c]{180}{$\Lsh$}
Clusters? Families\newline
\hspace*{1.35cm}\rotatebox[origin=c]{180}{$\Lsh$}Family studies\newline
\hspace*{2.1cm}\rotatebox[origin=c]{180}{$\Lsh$}
\textcolor{beamer@UIUCblue}{Twins data}
\end{column}
\end{columns}

\begin{center}
within-cluster dependence \(\Rightarrow\) within-family dependence
\end{center}

within-family dependence may reflect both disease \textbf{heritability}
and\newline the impact of shared \textbf{environmental effects}

###

\begin{minipage}{.15\textwidth}
 \begin{shaded}
  References
 \end{shaded}
\end{minipage}

+ \hypertarget{SCHEIKE}{
   [\textcolor{beamer@UIUCblue}{Cederkvist et al., 2019}]\newline
   L. Cederkvist, K. K. Holst, K. K. Andersen and T. H. Scheike (2019).
   Modeling the cumulative incidence function of multivariate competing
   risks data allowing for within-cluster dependence of risk and timing.
   \textit{Biostatistics}, v. 20, n. 2, p. 199-217}

<!-- ### -->

<!-- \begin{center} -->
<!--  \includegraphics[width=.9\linewidth]{thank_you.jpg} -->
<!-- \end{center} -->
